---
title: "New analyses"
author: "Daniel Aguirre"
date: "2026-06-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(phyloseq)
library(dplyr)
library(tibble)

setwd("C:/Users/DA.5015582/OneDrive - UAM/TRABAJO/Abiertos/Learning/Assembly")

ps <- readRDS("ps_nocontrol_filtrado_rare.rds")
metadata <- data.frame(sample_data(ps))
otu_mat <- as(otu_table(ps), "matrix")
if (!taxa_are_rows(ps)) otu_mat <- t(otu_mat)
taxonomy <- as.data.frame(tax_table(ps))

asv_prevalence <- rowSums(otu_mat > 0)
asv_total_abundance <- rowSums(otu_mat)
```

```{r create-asv-context-table, message=FALSE, warning=FALSE}

library(tidyr)

## Extract metadata
metadata <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID") %>%
  select(SampleID, Componentes)

## Extract ASV table
otu_mat <- as(otu_table(ps), "matrix")

if (!taxa_are_rows(ps)) {
  otu_mat <- t(otu_mat)
}

## Convert counts to relative abundance
## Since all samples are rarefied to the same depth, this is counts / 23394
otu_rel <- sweep(otu_mat, 2, colSums(otu_mat), "/")

## Long ASV table: one row per ASV x sample
asv_long <- as.data.frame(otu_rel) %>%
  rownames_to_column("ASV") %>%
  pivot_longer(
    cols = -ASV,
    names_to = "SampleID",
    values_to = "rel_abundance"
  ) %>%
  left_join(metadata, by = "SampleID")

## Summary per ASV x inoculum context
asv_context_summary <- asv_long %>%
  group_by(ASV, Componentes) %>%
  summarise(
    n_samples = n(),
    prevalence = sum(rel_abundance > 0),
    prevalence_prop = prevalence / n_samples,
    mean_abundance = mean(rel_abundance),
    median_abundance = median(rel_abundance),
    max_abundance = max(rel_abundance),
    .groups = "drop"
  )

## Add taxonomy
taxonomy <- as.data.frame(tax_table(ps)) %>%
  rownames_to_column("ASV")

asv_context_summary_tax <- asv_context_summary %>%
  left_join(taxonomy, by = "ASV")

## Quick checks
dim(asv_long)
dim(asv_context_summary_tax)

print(head(asv_long, 10))
print(head(asv_context_summary_tax, 10))

metadata_check <- data.frame(sample_data(ps)) |>
  tibble::rownames_to_column("SampleID")

table(metadata_check$Componentes)

asv_context_summary_tax |>
  dplyr::filter(mean_abundance > 0) |>
  dplyr::arrange(desc(mean_abundance)) |>
  print(n = 20)

## Save tables
write.csv(asv_long, "asv_long_relative_abundance.csv", row.names = FALSE)
write.csv(asv_context_summary_tax, "asv_context_summary_tax.csv", row.names = FALSE)

```

```{r}

## Filter ASVs already present in asv_long / asv_context_summary_tax
## Criterion: retain ASVs reaching >= 0.1% relative abundance in at least 2 samples.
## Rationale: this removes very rare/spurious ASVs while keeping taxa with evidence of reproducible ecological establishment.

#RESULT= 423 ASVs
asv_filter_summary <- asv_long %>%
  group_by(ASV) %>%
  summarise(
    n_samples_above_0.1percent = sum(rel_abundance >= 0.001),
    max_rel_abundance = max(rel_abundance),
    .groups = "drop"
  )

asvs_kept <- asv_filter_summary %>%
  filter(n_samples_above_0.1percent >= 2) %>%
  pull(ASV)

asv_long_filtered <- asv_long %>%
  filter(ASV %in% asvs_kept)

asv_context_summary_tax_filtered <- asv_context_summary_tax %>%
  filter(ASV %in% asvs_kept)

cat("Original ASVs:", length(unique(asv_long$ASV)), "\n")
cat("Filtered ASVs:", length(unique(asv_long_filtered$ASV)), "\n")

write.csv(asv_long_filtered,
          "asv_long_relative_abundance_filtered.csv",
          row.names = FALSE)

write.csv(asv_context_summary_tax_filtered,
          "asv_context_summary_tax_filtered.csv",
          row.names = FALSE)

```

```{r rank-abundance-global}
## ============================================================
## Rank-abundance analyses using the unfiltered ASV table
## Input object: asv_long
## ============================================================

library(dplyr)
library(ggplot2)
library(cowplot)

## ------------------------------------------------------------
## 1. Global rank-abundance curves, all samples together
## ------------------------------------------------------------

rank_abundance <- asv_long %>%
  filter(rel_abundance > 0) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()

p_rank_all <- ggplot(rank_abundance,
                     aes(x = rank,
                         y = rel_abundance,
                         group = SampleID)) +
  geom_line(alpha = 0.05, linewidth = 0.3) +
  scale_y_log10() +
  labs(
    x = "Rank",
    y = "Relative abundance",
    title = "Global rank-abundance distribution"
  ) +
  theme_bw()

p_rank_all

ggsave(
  "rank_abundance_all_unfiltered.png",
  plot = p_rank_all,
  width = 8,
  height = 6,
  dpi = 300
)


## ------------------------------------------------------------
## 2. Global rank-abundance envelope, all samples together
## ------------------------------------------------------------

rank_summary <- rank_abundance %>%
  group_by(rank) %>%
  summarise(
    q05 = quantile(rel_abundance, 0.05, na.rm = TRUE),
    q25 = quantile(rel_abundance, 0.25, na.rm = TRUE),
    median = median(rel_abundance, na.rm = TRUE),
    q75 = quantile(rel_abundance, 0.75, na.rm = TRUE),
    q95 = quantile(rel_abundance, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

p_envelope <- ggplot(rank_summary, aes(x = rank)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.15) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.3) +
  geom_line(aes(y = median), linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Rank",
    y = "Relative abundance",
    title = "Rank-abundance envelope across all samples"
  ) +
  theme_bw()

p_envelope

ggsave(
  "rank_abundance_all_envelope_unfiltered.png",
  plot = p_envelope,
  width = 8,
  height = 6,
  dpi = 300
)


## ------------------------------------------------------------
## 3. Rank-abundance envelopes by inoculum complexity
## ------------------------------------------------------------

rank_abundance_complexity <- asv_long %>%
  filter(rel_abundance > 0) %>%
  mutate(
    Complexity = case_when(
      nchar(Componentes) == 1 ~ "Single",
      nchar(Componentes) == 2 ~ "Pair",
      nchar(Componentes) == 3 ~ "Triplet",
      TRUE ~ NA_character_
    ),
    Complexity = factor(Complexity, levels = c("Single", "Pair", "Triplet"))
  ) %>%
  filter(!is.na(Complexity)) %>%
  group_by(SampleID, Complexity) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()

rank_summary_complexity <- rank_abundance_complexity %>%
  group_by(Complexity, rank) %>%
  summarise(
    q05 = quantile(rel_abundance, 0.05, na.rm = TRUE),
    q25 = quantile(rel_abundance, 0.25, na.rm = TRUE),
    median = median(rel_abundance, na.rm = TRUE),
    q75 = quantile(rel_abundance, 0.75, na.rm = TRUE),
    q95 = quantile(rel_abundance, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

p_envelope_complexity <- ggplot(rank_summary_complexity, aes(x = rank)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.15) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.3) +
  geom_line(aes(y = median), linewidth = 0.8) +
  scale_y_log10() +
  facet_grid(Complexity ~ ., scales = "fixed") +
  labs(
    x = "Rank",
    y = "Relative abundance",
    title = "Rank-abundance envelopes by inoculum complexity"
  ) +
  theme_bw()


p_envelope_complexity

ggsave(
  "rank_abundance_envelope_by_complexity_unfiltered.png",
  plot = p_envelope_complexity,
  width = 8,
  height = 10,
  dpi = 300
)


## ------------------------------------------------------------
## 4. Overlapping envelopes by inoculum complexity
## ------------------------------------------------------------
# 
# p_envelope_overlap <- ggplot(rank_summary_complexity,
#                              aes(x = rank,
#                                  color = Complexity,
#                                  fill = Complexity)) +
#   geom_ribbon(aes(ymin = q05, ymax = q95),
#               alpha = 0.12,
#               color = NA) +
#   geom_ribbon(aes(ymin = q25, ymax = q75),
#               alpha = 0.22,
#               color = NA) +
#   geom_line(aes(y = median),
#             linewidth = 0.9) +
#   scale_y_log10() +
#   scale_color_manual(values = c(
#     "Single" = "#1b9e77",
#     "Pair" = "#d95f02",
#     "Triplet" = "#7570b3"
#   )) +
#   scale_fill_manual(values = c(
#     "Single" = "#1b9e77",
#     "Pair" = "#d95f02",
#     "Triplet" = "#7570b3"
#   )) +
#   labs(
#     x = "Rank",
#     y = "Relative abundance",
#     title = "Overlapping rank-abundance envelopes by inoculum complexity",
#     color = "Complexity",
#     fill = "Complexity"
#   ) +
#   theme_bw()
# 
# p_envelope_overlap
# 
# ggsave(
#   "rank_abundance_envelope_complexity_overlap_unfiltered.png",
#   plot = p_envelope_overlap,
#   width = 8,
#   height = 6,
#   dpi = 300
# )




p_envelope_overlap <- ggplot(rank_summary_complexity,
                             aes(x = rank,
                                 color = Complexity,
                                 fill = Complexity)) +
  geom_ribbon(aes(ymin = q05, ymax = q95),
              alpha = 0.12,
              color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75),
              alpha = 0.22,
              color = NA) +
  geom_line(aes(y = median),
            linewidth = 0.9) +
  scale_y_log10() +
  scale_color_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  scale_fill_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  labs(
    x = "Rank",
    y = "Relative abundance",
    title = NULL,
    color = "Complexity",
    fill = "Complexity"
  ) +
  theme_bw() +
  theme(
    legend.position = c(0.66, 0.35),
    legend.background = element_rect(
      fill = scales::alpha("white", 0.75),
      colour = NA
    )
  )

## ------------------------------------------------------------
## 5. Inset plot: same analysis but only ASVs >1%
## ------------------------------------------------------------

rank_abundance_gt1 <- asv_long %>%
  filter(rel_abundance > 0.01) %>%
  mutate(
    Complexity = case_when(
      nchar(Componentes) == 1 ~ "Single",
      nchar(Componentes) == 2 ~ "Pair",
      nchar(Componentes) == 3 ~ "Triplet",
      TRUE ~ NA_character_
    ),
    Complexity = factor(Complexity, levels = c("Single", "Pair", "Triplet"))
  ) %>%
  filter(!is.na(Complexity)) %>%
  group_by(SampleID, Complexity) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()

rank_summary_gt1 <- rank_abundance_gt1 %>%
  group_by(Complexity, rank) %>%
  summarise(
    q05 = quantile(rel_abundance, 0.05, na.rm = TRUE),
    q25 = quantile(rel_abundance, 0.25, na.rm = TRUE),
    median = median(rel_abundance, na.rm = TRUE),
    q75 = quantile(rel_abundance, 0.75, na.rm = TRUE),
    q95 = quantile(rel_abundance, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

p_inset <- ggplot(rank_summary_gt1,
                  aes(x = rank,
                      color = Complexity,
                      fill = Complexity)) +
  geom_ribbon(aes(ymin = q05, ymax = q95),
              alpha = 0.12,
              color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75),
              alpha = 0.22,
              color = NA) +
  geom_line(aes(y = median), linewidth = 0.7) +
  scale_y_log10() +
  scale_color_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  scale_fill_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  labs(
    x = "Rank",
    y = NULL,
    title = ">1%"
  ) +
  theme_bw(base_size = 8) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 9),
    axis.title.x = element_text(size = 8),
    axis.text = element_text(size = 7)
  )

p_combined <- ggdraw() +
  draw_plot(p_envelope_overlap) +
  draw_plot(
    p_inset,
    x = 0.45,
    y = 0.52,
    width = 0.40,
    height = 0.40
  )

p_combined

ggsave(
  "rank_abundance_envelope_complexity_overlap_with_gt1_inset_unfiltered.png",
  plot = p_combined,
  width = 8,
  height = 6,
  dpi = 300
)








## ------------------------------------------------------------
## 6. Cumulative abundance: ASVs needed to reach 50% and 90%
## ------------------------------------------------------------

cumulative_rank_abundance <- asv_long %>%
  filter(rel_abundance > 0) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  mutate(
    rank = row_number(),
    cumulative_abundance = cumsum(rel_abundance)
  ) %>%
  ungroup()

n_asvs_summary <- cumulative_rank_abundance %>%
  group_by(SampleID) %>%
  summarise(
    n_asvs_50 = min(rank[cumulative_abundance >= 0.50]),
    n_asvs_90 = min(rank[cumulative_abundance >= 0.90]),
    .groups = "drop"
  )

summary_stats <- n_asvs_summary %>%
  summarise(
    median_50 = median(n_asvs_50, na.rm = TRUE),
    q25_50 = quantile(n_asvs_50, 0.25, na.rm = TRUE),
    q75_50 = quantile(n_asvs_50, 0.75, na.rm = TRUE),
    median_90 = median(n_asvs_90, na.rm = TRUE),
    q25_90 = quantile(n_asvs_90, 0.25, na.rm = TRUE),
    q75_90 = quantile(n_asvs_90, 0.75, na.rm = TRUE)
  )

print(summary_stats)
```

```{r bray-nmds-pcoa, message=FALSE, warning=FALSE}

set.seed(123)

## ============================================================
## 1. NMDS with all samples
##    Bray-Curtis, labels only, no points
## ============================================================

ord_nmds_bray_all <- ordinate(
  ps,
  method = "NMDS",
  distance = "bray",
  trymax = 100
)


## ============================================================
## 2. NMDS using only single-inoculum samples
##    Important: subset first, then run NMDS on this subset
## ============================================================

ps_singles <- subset_samples(ps, nchar(Componentes) == 1)

## Optional: remove ASVs absent from the singles subset
ps_singles <- prune_taxa(taxa_sums(ps_singles) > 0, ps_singles)

ord_nmds_bray_singles <- ordinate(
  ps_singles,
  method = "NMDS",
  distance = "bray",
  trymax = 100
)



ord_df_all <- plot_ordination(ps, ord_nmds_bray_all, type = "samples", justDF = TRUE)

p_nmds_bray_all <- ggplot(ord_df_all, aes(x = NMDS1, y = NMDS2)) +
  geom_text(aes(label = Componentes), size = 3, check_overlap = TRUE) +
  labs(title = "NMDS - Bray-Curtis, all samples") +
  theme_bw()

ggsave(
  "NMDS_Bray_Curtis_all_labels_only.png",
  plot = p_nmds_bray_all,
  width = 8,
  height = 6,
  dpi = 300
)

ord_df_singles <- plot_ordination(ps_singles, ord_nmds_bray_singles,
                                  type = "samples", justDF = TRUE)

p_nmds_bray_singles <- ggplot(ord_df_singles,
                              aes(x = NMDS1, y = NMDS2)) +
  geom_text(aes(label = Componentes, color = Componentes),
            size = 4, check_overlap = FALSE) +
  labs(title = "NMDS - Bray-Curtis, single inocula only",
       color = "Component") +
  theme_bw()

ggsave(
  "NMDS_Bray_Curtis_singles_labels_only.png",
  plot = p_nmds_bray_singles,
  width = 8,
  height = 6,
  dpi = 300
)


pA <- p_nmds_bray_singles +
  labs(title = NULL, color = NULL) +
  theme(legend.position = "none") +
  annotate("text", x = -Inf, y = Inf, label = "A",
           hjust = -0.7, vjust = 1.5, fontface = "bold", size = 6)

pB <- p_nmds_bray_all +
  labs(title = NULL) +
  theme(legend.position = "none") +
  annotate("text", x = -Inf, y = Inf, label = "B",
           hjust = -0.7, vjust = 1.5, fontface = "bold", size = 6)

p_nmds_combined <- cowplot::plot_grid(
  pA, pB,
  nrow = 1,
  labels = NULL,
  align = "hv",
  rel_widths = c(1, 1.25)
)

ggsave(
  "NMDS_Bray_Curtis_singles_and_all.png",
  plot = p_nmds_combined,
  width = 12,
  height = 5,
  dpi = 300
)








## ============================================================
## PCoA using the same phyloseq objects and Bray–Curtis
## dissimilarity used for the NMDS ordinations
## ============================================================

## Figure 5B: all samples
ord_pcoa_bray_all <- ordinate(
  ps,
  method = "PCoA",
  distance = "bray"
)

## Figure 5A: single-inoculum samples
ord_pcoa_bray_singles <- ordinate(
  ps_singles,
  method = "PCoA",
  distance = "bray"
)

## ============================================================
## Extract PCoA coordinates
## ============================================================

pcoa_df_all <- plot_ordination(
  ps,
  ord_pcoa_bray_all,
  type = "samples",
  justDF = TRUE
)

pcoa_df_singles <- plot_ordination(
  ps_singles,
  ord_pcoa_bray_singles,
  type = "samples",
  justDF = TRUE
)

## Check coordinate names
colnames(pcoa_df_all)
colnames(pcoa_df_singles)

## In phyloseq PCoA outputs, the first coordinates are normally
## called Axis.1 and Axis.2. Rename them for plotting.
names(pcoa_df_all)[
  names(pcoa_df_all) == "Axis.1"
] <- "PCoA1"

names(pcoa_df_all)[
  names(pcoa_df_all) == "Axis.2"
] <- "PCoA2"

names(pcoa_df_singles)[
  names(pcoa_df_singles) == "Axis.1"
] <- "PCoA1"

names(pcoa_df_singles)[
  names(pcoa_df_singles) == "Axis.2"
] <- "PCoA2"

## ============================================================
## Percentages represented by PCoA axes
## ============================================================

variance_all <- 100 *
  ord_pcoa_bray_all$values$Relative_eig[1:2]

variance_singles <- 100 *
  ord_pcoa_bray_singles$values$Relative_eig[1:2]

pcoa_axis_percentages <- data.frame(
  Panel = c(
    "Figure 5A: single-inoculum communities",
    "Figure 5B: all communities"
  ),
  PCoA1_percent = c(
    variance_singles[1],
    variance_all[1]
  ),
  PCoA2_percent = c(
    variance_singles[2],
    variance_all[2]
  )
)

print(pcoa_axis_percentages)

write.csv(
  pcoa_axis_percentages,
  "Figure5_PCoA_axis_percentages.csv",
  row.names = FALSE
)

## ============================================================
## Figure 5B equivalent: all samples
## ============================================================

p_pcoa_bray_all <- ggplot(
  pcoa_df_all,
  aes(x = PCoA1, y = PCoA2)
) +
  geom_text(
    aes(label = Componentes),
    size = 3,
    check_overlap = FALSE
  ) +
  labs(
    title = "PCoA – Bray–Curtis, all samples",
    x = paste0(
      "PCoA1 (",
      round(variance_all[1], 1),
      "%)"
    ),
    y = paste0(
      "PCoA2 (",
      round(variance_all[2], 1),
      "%)"
    )
  ) +
  theme_bw()

ggsave(
  "PCoA_Bray_Curtis_all_labels_only.png",
  plot = p_pcoa_bray_all,
  width = 8,
  height = 6,
  dpi = 300
)

## ============================================================
## Figure 5A equivalent: single-inoculum samples
## ============================================================

p_pcoa_bray_singles <- ggplot(
  pcoa_df_singles,
  aes(x = PCoA1, y = PCoA2)
) +
  geom_text(
    aes(
      label = Componentes,
      color = Componentes
    ),
    size = 4,
    check_overlap = FALSE
  ) +
  labs(
    title = "PCoA – Bray–Curtis, single inocula only",
    x = paste0(
      "PCoA1 (",
      round(variance_singles[1], 1),
      "%)"
    ),
    y = paste0(
      "PCoA2 (",
      round(variance_singles[2], 1),
      "%)"
    ),
    color = "Component"
  ) +
  theme_bw()

ggsave(
  "PCoA_Bray_Curtis_singles_labels_only.png",
  plot = p_pcoa_bray_singles,
  width = 8,
  height = 6,
  dpi = 300
)

## ============================================================
## Combined PCoA figure
## ============================================================

pA_pcoa <- p_pcoa_bray_singles +
  labs(title = NULL, color = NULL) +
  theme(legend.position = "none") +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    label = "A",
    hjust = -0.7,
    vjust = 1.5,
    fontface = "bold",
    size = 6
  )

pB_pcoa <- p_pcoa_bray_all +
  labs(title = NULL) +
  theme(legend.position = "none") +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    label = "B",
    hjust = -0.7,
    vjust = 1.5,
    fontface = "bold",
    size = 6
  )

p_pcoa_combined <- cowplot::plot_grid(
  pA_pcoa,
  pB_pcoa,
  nrow = 1,
  labels = NULL,
  align = "hv",
  rel_widths = c(1, 1.25)
)

p_pcoa_combined

ggsave(
  "PCoA_Bray_Curtis_singles_and_all.png",
  plot = p_pcoa_combined,
  width = 12,
  height = 5,
  dpi = 300
)
```

```{r, message=FALSE, warning=FALSE}
top2_pair_freq <- asv_long %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  filter(rel_abundance > 0) %>%
  mutate(Genus = ifelse(is.na(Genus), "Unassigned", Genus)) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  select(SampleID, rank, Genus) %>%
  tidyr::pivot_wider(
    names_from = rank,
    values_from = Genus,
    names_prefix = "rank"
  ) %>%
  mutate(pair = paste(rank1, rank2, sep = " - ")) %>%
  count(pair, name = "n_samples") %>%
  mutate(percent = 100 * n_samples / sum(n_samples)) %>%
  arrange(desc(percent))

print(top2_pair_freq, n = 20)

#No Pseudomonas domination:

non_pseudomonas_top2 <- asv_long %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  filter(rel_abundance > 0) %>%
  mutate(Genus = ifelse(is.na(Genus), "Unassigned", Genus)) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  summarise(
    top1 = Genus[1],
    top2 = Genus[2],
    Componentes = first(Componentes),
    both_non_pseudomonas = all(Genus != "Pseudomonas"),
    .groups = "drop"
  ) %>%
  filter(both_non_pseudomonas)

print(non_pseudomonas_top2, n = Inf)

top2_ACD <- asv_long %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  filter(rel_abundance > 0) %>%
  filter(Componentes %in% c("A", "C", "D")) %>%
  mutate(Genus = ifelse(is.na(Genus), "Unassigned", Genus)) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  mutate(rank = row_number()) %>%
  ungroup()

print(top2_ACD, n = Inf)



asv_ids <- asv_long %>%
  group_by(ASV) %>%
  summarise(global_abundance = sum(rel_abundance), .groups = "drop") %>%
  arrange(desc(global_abundance)) %>%
  mutate(ASV_ID = paste0("ASV_", row_number()))

print(head(asv_ids, 20))



top2_asvs_global <- asv_long %>%
  left_join(asv_ids, by = "ASV") %>%
  filter(rel_abundance > 0) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  count(rank, ASV_ID, name = "n_samples") %>%
  group_by(rank) %>%
  mutate(percent = 100 * n_samples / sum(n_samples)) %>%
  arrange(rank, desc(percent))

print(top2_asvs_global, n = 20)


top2_selected <- asv_long %>%
  left_join(asv_ids, by = "ASV") %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  filter(rel_abundance > 0) %>%
  filter(Componentes %in% c("G", "A", "C", "D", "CG", "ADG")) %>%
  mutate(Genus = ifelse(is.na(Genus), "Unassigned", Genus)) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  select(SampleID, Componentes, rank, ASV_ID, Genus, rel_abundance)

top2_selected_wide <- top2_selected %>%
  tidyr::pivot_wider(
    names_from = rank,
    values_from = c(ASV_ID, Genus, rel_abundance),
    names_glue = "{.value}_rank{rank}"
  ) %>%
  arrange(Componentes)

print(top2_selected_wide, n = Inf)


asv_focus_singles <- asv_long %>%
  left_join(asv_ids, by = "ASV") %>%
  filter(nchar(Componentes) == 1) %>%
  filter(ASV_ID %in% c("ASV_1", "ASV_2", "ASV_9", "ASV_10")) %>%
  select(SampleID, Componentes, ASV_ID, rel_abundance)

print(asv_focus_singles, n = Inf)

ggplot(asv_focus_singles,
       aes(x = Componentes,
           y = rel_abundance,
           fill = ASV_ID)) +
  geom_boxplot() +
  facet_wrap(~ASV_ID, scales = "free_y") +
  theme_bw()


pseudomonas_abundance_top10 <- asv_long %>%
  group_by(ASV) %>%
  summarise(mean_rel_abundance = mean(rel_abundance), .groups = "drop") %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  filter(Genus == "Pseudomonas") %>%
  left_join(asv_ids, by = "ASV") %>%
  arrange(as.numeric(gsub("ASV_", "", ASV_ID))) %>%
  slice_head(n = 10)

p_pseudomonas_top10 <- ggplot(
  pseudomonas_abundance_top10,
  aes(
    x = reorder(ASV_ID, as.numeric(gsub("ASV_", "", ASV_ID))),
    y = mean_rel_abundance
  )
) +
  geom_col() +
  geom_text(
    aes(label = round(mean_rel_abundance, 3)),
    vjust = -0.3,
    size = 3
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    x = "Pseudomonas ASV",
    y = "Mean relative abundance",
    title = "Top 10 Pseudomonas ASVs (out of 186 total)"
  )

p_pseudomonas_top10

ggsave(
  "top10_pseudomonas_asvs_mean_relative_abundance.png",
  plot = p_pseudomonas_top10,
  width = 8,
  height = 5,
  dpi = 300
)
```

#new abundant ASVs analysis, uses objects from previous chunks
```{r dominant-asv-abundance-across-complexity, message=FALSE, warning=FALSE}

library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

## ------------------------------------------------------------
## 1. Focal dominant ASVs
## ------------------------------------------------------------

focal_asvs <- c("ASV_1", "ASV_2", "ASV_9", "ASV_10")

## Taxonomic labels for the figure
focal_taxonomy <- asv_context_summary_tax %>%
  select(ASV, Genus) %>%
  distinct() %>%
  left_join(asv_ids, by = "ASV") %>%
  filter(ASV_ID %in% focal_asvs) %>%
  mutate(
    ASV_number = as.numeric(sub("ASV_", "", ASV_ID))
  ) %>%
  arrange(ASV_number) %>%
  mutate(
    ASV_label = paste0(ASV_ID, " (", Genus, ")")
  ) %>%
  select(ASV, ASV_ID, Genus, ASV_label)

print(focal_taxonomy)

## ------------------------------------------------------------
## 2. Define the source communities of each focal ASV
## ------------------------------------------------------------

## A single-inoculum community is considered a source when the
## ASV has a mean relative abundance of at least 0.1% across its
## replicate samples. This follows the source-definition criterion
## used elsewhere in the manuscript.

focal_source_communities <- asv_long %>%
  left_join(asv_ids, by = "ASV") %>%
  filter(
    ASV_ID %in% focal_asvs,
    nchar(Componentes) == 1
  ) %>%
  group_by(ASV_ID, Componentes) %>%
  summarise(
    mean_single_abundance = mean(rel_abundance),
    .groups = "drop"
  ) %>%
  filter(mean_single_abundance >= 0.001) %>%
  group_by(ASV_ID) %>%
  summarise(
    source_communities = paste(Componentes, collapse = ""),
    source_list = list(Componentes),
    .groups = "drop"
  )

print(focal_source_communities)

## ------------------------------------------------------------
## 3. Abundance in eligible inoculum combinations
## ------------------------------------------------------------

## A combination is eligible when it contains at least one source
## community in which the focal ASV met the criterion above.

focal_asv_eligible_samples <- asv_long %>%
  left_join(asv_ids, by = "ASV") %>%
  filter(ASV_ID %in% focal_asvs) %>%
  left_join(focal_source_communities, by = "ASV_ID") %>%
  rowwise() %>%
  mutate(
    eligible = any(
      unlist(source_list) %in%
        strsplit(as.character(Componentes), "")[[1]]
    )
  ) %>%
  ungroup() %>%
  filter(eligible) %>%
  mutate(
    Complexity = case_when(
      nchar(Componentes) == 1 ~ "Single",
      nchar(Componentes) == 2 ~ "Pair",
      nchar(Componentes) == 3 ~ "Triplet",
      TRUE ~ NA_character_
    ),
    Complexity = factor(
      Complexity,
      levels = c("Single", "Pair", "Triplet")
    )
  ) %>%
  filter(!is.na(Complexity))

## ------------------------------------------------------------
## 4. Average replicates within each inoculum combination
## ------------------------------------------------------------

focal_asv_treatment_means <- focal_asv_eligible_samples %>%
  group_by(
    ASV,
    ASV_ID,
    Componentes,
    Complexity
  ) %>%
  summarise(
    n_replicates = n(),
    mean_relative_abundance = mean(rel_abundance),
    sd_relative_abundance = sd(rel_abundance),
    .groups = "drop"
  ) %>%
  left_join(
    focal_taxonomy %>%
      select(ASV, ASV_ID, Genus, ASV_label),
    by = c("ASV", "ASV_ID")
  ) %>%
  mutate(
    ASV_label = factor(
      ASV_label,
      levels = focal_taxonomy$ASV_label
    )
  )

print(
  focal_asv_treatment_means %>%
    arrange(ASV_ID, Complexity, Componentes),
  n = Inf
)

## ------------------------------------------------------------
## 5. Descriptive summary by complexity class
## ------------------------------------------------------------

focal_asv_complexity_summary <- focal_asv_treatment_means %>%
  group_by(
    ASV_ID,
    Genus,
    ASV_label,
    Complexity
  ) %>%
  summarise(
    n_combinations = n(),
    median_abundance = median(mean_relative_abundance),
    q1_abundance = quantile(mean_relative_abundance, 0.25),
    q3_abundance = quantile(mean_relative_abundance, 0.75),
    mean_abundance = mean(mean_relative_abundance),
    .groups = "drop"
  )

print(focal_asv_complexity_summary, n = Inf)

## ------------------------------------------------------------
## 6. Supplementary Figure S25A
## ------------------------------------------------------------

complexity_colors <- c(
  "Single"  = "#1b9e77",
  "Pair"    = "#d95f02",
  "Triplet" = "#7570b3"
)

p_dominant_asv_complexity <- ggplot(
  focal_asv_treatment_means,
  aes(
    x = Complexity,
    y = mean_relative_abundance,
    fill = Complexity,
    color = Complexity
  )
) +
  geom_boxplot(
    width = 0.60,
    alpha = 0.25,
    outlier.shape = NA
  ) +
  geom_jitter(
    width = 0.12,
    height = 0,
    size = 2,
    alpha = 0.80
  ) +
  facet_wrap(
    ~ ASV_label,
    nrow = 1,
    scales = "free_y"
  ) +
  scale_fill_manual(
    values = complexity_colors,
    drop = FALSE
  ) +
  scale_color_manual(
    values = complexity_colors,
    drop = FALSE
  ) +
  scale_y_continuous(
    labels = label_percent(accuracy = 1)
  ) +
  labs(
    x = "Inoculum complexity",
    y = "Mean relative abundance per inoculum combination",
    fill = "Complexity",
    color = "Complexity"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(face = "bold")
  )

p_dominant_asv_complexity

ggsave(
  "Figure_S25A_dominant_ASVs_across_inoculum_complexity.png",
  plot = p_dominant_asv_complexity,
  width = 12,
  height = 4,
  dpi = 300
)

## ------------------------------------------------------------
## 7. Save underlying values
## ------------------------------------------------------------

write.csv(
  focal_asv_treatment_means,
  "Figure_S25A_dominant_ASV_treatment_means.csv",
  row.names = FALSE
)

write.csv(
  focal_asv_complexity_summary,
  "Figure_S25A_dominant_ASV_summary_by_complexity.csv",
  row.names = FALSE
)

write.csv(
  focal_source_communities %>%
    select(-source_list),
  "Figure_S25A_dominant_ASV_source_communities.csv",
  row.names = FALSE
)
```

```

#4-panel nmds colored by abundant ASV

```{r dominant-asv-abundance-global-nmds, message=FALSE, warning=FALSE}

library(dplyr)
library(tibble)
library(ggplot2)
library(scales)
library(viridis)

## ------------------------------------------------------------
## 1. Recover the coordinates of the global NMDS used in Fig. 5B
## ------------------------------------------------------------

nmds_coordinates_all <- ord_df_all

## plot_ordination() normally retains sample names as row names.
## Convert them explicitly into a SampleID column if necessary.

if (!"SampleID" %in% colnames(nmds_coordinates_all)) {
  nmds_coordinates_all <- nmds_coordinates_all %>%
    rownames_to_column("SampleID")
}

## Check that every sample in the NMDS has a unique identifier
stopifnot(
  !anyDuplicated(nmds_coordinates_all$SampleID)
)

## ------------------------------------------------------------
## 2. Abundance of each focal ASV in every sample
## ------------------------------------------------------------

focal_asv_nmds_data <- asv_long %>%
  left_join(
    asv_ids,
    by = "ASV"
  ) %>%
  filter(
    ASV_ID %in% focal_asvs
  ) %>%
  left_join(
    focal_taxonomy %>%
      select(ASV, ASV_ID, Genus, ASV_label),
    by = c("ASV", "ASV_ID")
  ) %>%
  select(
    SampleID,
    Componentes,
    ASV_ID,
    Genus,
    ASV_label,
    rel_abundance
  ) %>%
  left_join(
    nmds_coordinates_all %>%
      select(
        SampleID,
        NMDS1,
        NMDS2
      ),
    by = "SampleID"
  ) %>%
  mutate(
    ASV_number = as.numeric(
      sub("ASV_", "", ASV_ID)
    ),
    ASV_label = factor(
      ASV_label,
      levels = focal_taxonomy$ASV_label
    )
  )

## ------------------------------------------------------------
## 3. Quality checks
## ------------------------------------------------------------

## Every focal ASV should have one value for every NMDS sample
print(
  focal_asv_nmds_data %>%
    count(ASV_ID, name = "n_samples")
)

print(
  paste(
    "Number of samples in global NMDS:",
    nrow(nmds_coordinates_all)
  )
)

## No missing NMDS coordinates should remain
stopifnot(
  !anyNA(focal_asv_nmds_data$NMDS1),
  !anyNA(focal_asv_nmds_data$NMDS2)
)

## Inspect abundance ranges
print(
  focal_asv_nmds_data %>%
    group_by(ASV_ID, Genus) %>%
    summarise(
      minimum = min(rel_abundance),
      median = median(rel_abundance),
      maximum = max(rel_abundance),
      .groups = "drop"
    )
)

## ------------------------------------------------------------
## 4. Supplementary Figure S25B: 2 × 2 panels
##    Independent color scale and legend for each ASV
## ------------------------------------------------------------

library(patchwork)

## Preserve the desired ASV order
focal_asv_order <- c(
  "ASV_1",
  "ASV_2",
  "ASV_9",
  "ASV_10"
)

## Function generating one NMDS panel per ASV
make_asv_nmds_panel <- function(current_asv) {

  panel_data <- focal_asv_nmds_data %>%
    filter(ASV_ID == current_asv)

  panel_title <- unique(
    as.character(panel_data$ASV_label)
  )

  ggplot(
    panel_data,
    aes(
      x = NMDS1,
      y = NMDS2,
      color = rel_abundance
    )
  ) +
    geom_point(
      size = 2.5,
      alpha = 0.90
    ) +
    scale_color_viridis_c(
      option = "magma",
      direction = -1,
      trans = "sqrt",
      limits = range(
        panel_data$rel_abundance,
        na.rm = TRUE
      ),
      labels = label_percent(accuracy = 1),
      name = "Relative\nabundance"
    ) +
    coord_equal() +
    labs(
      title = panel_title,
      x = "NMDS1",
      y = "NMDS2"
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(
        face = "bold",
        hjust = 0.5,
        size = 11
      ),
      legend.position = "right",
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 7)
    )
}

## Generate the four independent plots
dominant_asv_nmds_panels <- lapply(
  focal_asv_order,
  make_asv_nmds_panel
)

## Combine as a 2 × 2 figure.
## Each panel retains its own abundance scale and legend.
p_dominant_asv_nmds <- wrap_plots(
  dominant_asv_nmds_panels,
  ncol = 2,
  nrow = 2,
  guides = "keep"
) +
  plot_annotation(
    tag_levels = "A"
  )

p_dominant_asv_nmds

ggsave(
  "Figure_S25B_dominant_ASV_abundance_global_NMDS.png",
  plot = p_dominant_asv_nmds,
  width = 11,
  height = 8,
  dpi = 300
)

## ------------------------------------------------------------
## 5. Save underlying data
## ------------------------------------------------------------

write.csv(
  focal_asv_nmds_data,
  "Figure_S25B_dominant_ASV_abundance_global_NMDS.csv",
  row.names = FALSE
)
```

#Abundant ASV analysis on dominance results

```{r}

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(scales)
library(patchwork)

## ============================================================
## 1. Qualitative outcomes used in the dominance network
## ============================================================

txt <- c(
  "A gana a C",
  "D gana a A",
  "E gana a A",
  "F gana a A",
  "A gana a G",
  "A empata con H",
  "D gana a C",
  "E gana a C",
  "F gana a C",
  "C empata con G",
  "H gana a C",
  "E gana a D",
  "D gana a F",
  "D gana a G",
  "D empata con H",
  "E gana a F",
  "E gana a G",
  "E gana a H",
  "F gana a G",
  "F empata con H",
  "H gana a G"
)

community_order <- c(
  "A", "C", "D", "E", "F", "G", "H"
)

parse_dominance_outcome <- function(current_text) {

  if (str_detect(current_text, "gana a")) {

    parsed <- str_match(
      current_text,
      "^([A-Z]) gana a ([A-Z])$"
    )

    winner <- parsed[, 2]
    loser  <- parsed[, 3]

    pair_members <- community_order[
      community_order %in% c(winner, loser)
    ]

    data.frame(
      Original_text = current_text,
      Pair = paste0(pair_members, collapse = ""),
      Community_1 = pair_members[1],
      Community_2 = pair_members[2],
      Outcome = "Dominance",
      Winner = winner,
      Loser = loser,
      stringsAsFactors = FALSE
    )

  } else if (str_detect(current_text, "empata con")) {

    parsed <- str_match(
      current_text,
      "^([A-Z]) empata con ([A-Z])$"
    )

    pair_members <- community_order[
      community_order %in% c(parsed[, 2], parsed[, 3])
    ]

    data.frame(
      Original_text = current_text,
      Pair = paste0(pair_members, collapse = ""),
      Community_1 = pair_members[1],
      Community_2 = pair_members[2],
      Outcome = "No clear dominance",
      Winner = NA_character_,
      Loser = NA_character_,
      stringsAsFactors = FALSE
    )

  } else {

    stop(
      paste(
        "Unrecognized dominance description:",
        current_text
      )
    )
  }
}

dominance_outcomes <- bind_rows(
  lapply(
    txt,
    parse_dominance_outcome
  )
) %>%
  mutate(
    Pair_order = row_number()
  )

print(dominance_outcomes, n = Inf)

stopifnot(
  nrow(dominance_outcomes) == 21,
  !anyDuplicated(dominance_outcomes$Pair)
)

## ============================================================
## 2. Mean abundance of focal ASVs in singles and pairs
## ============================================================

focal_asv_pair_means <- asv_long %>%
  left_join(
    asv_ids,
    by = "ASV"
  ) %>%
  filter(
    ASV_ID %in% focal_asvs,
    nchar(as.character(Componentes)) %in% c(1, 2)
  ) %>%
  group_by(
    ASV,
    ASV_ID,
    Componentes
  ) %>%
  summarise(
    n_replicates = n(),
    mean_relative_abundance = mean(
      rel_abundance,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

## ============================================================
## 3. One row per pair × focal ASV
## ============================================================

dominant_asv_pair_dynamics <- tidyr::crossing(
  dominance_outcomes,
  ASV_ID = focal_asvs
) %>%
  left_join(
    focal_asv_pair_means %>%
      select(
        ASV_ID,
        Componentes,
        Pair_abundance = mean_relative_abundance
      ),
    by = c(
      "ASV_ID",
      "Pair" = "Componentes"
    )
  ) %>%
  left_join(
    focal_asv_pair_means %>%
      select(
        ASV_ID,
        Componentes,
        Community_1_abundance = mean_relative_abundance
      ),
    by = c(
      "ASV_ID",
      "Community_1" = "Componentes"
    )
  ) %>%
  left_join(
    focal_asv_pair_means %>%
      select(
        ASV_ID,
        Componentes,
        Community_2_abundance = mean_relative_abundance
      ),
    by = c(
      "ASV_ID",
      "Community_2" = "Componentes"
    )
  ) %>%
  mutate(
    Source_mean_abundance =
      (
        Community_1_abundance +
        Community_2_abundance
      ) / 2,

    ## Change in the pair relative to the mean abundance
    ## observed in its two corresponding singles
    Delta_from_source_mean =
      Pair_abundance -
      Source_mean_abundance,

    Winner_abundance = case_when(
      Outcome == "Dominance" &
        Winner == Community_1 ~
        Community_1_abundance,

      Outcome == "Dominance" &
        Winner == Community_2 ~
        Community_2_abundance,

      TRUE ~ NA_real_
    ),

    Loser_abundance = case_when(
      Outcome == "Dominance" &
        Loser == Community_1 ~
        Community_1_abundance,

      Outcome == "Dominance" &
        Loser == Community_2 ~
        Community_2_abundance,

      TRUE ~ NA_real_
    ),

    Winner_minus_loser =
      Winner_abundance -
      Loser_abundance,

    Source_concordance = case_when(
      Outcome == "No clear dominance" ~
        "No clear dominance",

      Winner_abundance > Loser_abundance ~
        "Winner-enriched",

      Winner_abundance < Loser_abundance ~
        "Loser-enriched",

      Winner_abundance == Loser_abundance ~
        "Equal",

      TRUE ~ NA_character_
    ),

    Cell_label = case_when(
      Source_concordance == "Winner-enriched" ~ "W",
      Source_concordance == "Loser-enriched" ~ "L",
      Source_concordance == "Equal" ~ "=",
      Source_concordance == "No clear dominance" ~ "–",
      TRUE ~ ""
    )
  ) %>%
  left_join(
    focal_taxonomy %>%
      select(
        ASV_ID,
        Genus,
        ASV_label
      ) %>%
      distinct(),
    by = "ASV_ID"
  )

## Check for missing abundance values
stopifnot(
  !anyNA(
    dominant_asv_pair_dynamics$Pair_abundance
  ),
  !anyNA(
    dominant_asv_pair_dynamics$Community_1_abundance
  ),
  !anyNA(
    dominant_asv_pair_dynamics$Community_2_abundance
  )
)

print(
  dominant_asv_pair_dynamics %>%
    select(
      Pair,
      Outcome,
      Winner,
      Loser,
      ASV_ID,
      Community_1_abundance,
      Community_2_abundance,
      Pair_abundance,
      Delta_from_source_mean,
      Source_concordance
    ),
  n = Inf
)

## ============================================================
## 4. Descriptive concordance summary
## ============================================================

dominance_asv_concordance_summary <-
  dominant_asv_pair_dynamics %>%
  filter(
    Outcome == "Dominance"
  ) %>%
  group_by(
    ASV_ID,
    Genus
  ) %>%
  summarise(
    n_dominance_comparisons = n(),

    n_winner_enriched = sum(
      Source_concordance == "Winner-enriched"
    ),

    n_loser_enriched = sum(
      Source_concordance == "Loser-enriched"
    ),

    n_equal = sum(
      Source_concordance == "Equal"
    ),

    proportion_winner_enriched =
      n_winner_enriched /
      n_dominance_comparisons,

    median_winner_minus_loser =
      median(
        Winner_minus_loser,
        na.rm = TRUE
      ),

    median_delta_in_pair =
      median(
        Delta_from_source_mean,
        na.rm = TRUE
      ),

    .groups = "drop"
  )

print(
  dominance_asv_concordance_summary,
  n = Inf
)

## ============================================================
## 5. Labels and ordering for Figure S25C
## ============================================================

pair_plot_labels <- dominance_outcomes %>%
  mutate(
    Pair_label = case_when(
      Outcome == "Dominance" ~
        paste0(Pair, ": ", Winner, " > ", Loser),

      Outcome == "No clear dominance" ~
        paste0(Pair, ": no clear dominance")
    )
  ) %>%
  select(
    Pair,
    Pair_label,
    Pair_order
  )

dominant_asv_pair_dynamics <-
  dominant_asv_pair_dynamics %>%
  left_join(
    pair_plot_labels,
    by = c(
      "Pair",
      "Pair_order"
    )
  ) %>%
  mutate(
    Pair_label = factor(
      Pair_label,
      levels = rev(
        pair_plot_labels$Pair_label
      )
    )
  )

## ============================================================
## 6. Figure S25C
##    Independent heatmap scale for each ASV
## ============================================================

make_dominance_heatmap <- function(current_asv) {

  panel_data <- dominant_asv_pair_dynamics %>%
    filter(
      ASV_ID == current_asv
    )

  panel_limit <- max(
    abs(panel_data$Delta_from_source_mean),
    na.rm = TRUE
  )

  ## Avoid a zero-width color scale
  if (
    !is.finite(panel_limit) ||
    panel_limit == 0
  ) {
    panel_limit <- 0.001
  }

  panel_title <- unique(
    panel_data$ASV_label
  )

  ggplot(
    panel_data,
    aes(
      x = ASV_ID,
      y = Pair_label,
      fill = Delta_from_source_mean
    )
  ) +
    geom_tile(
      color = "white",
      linewidth = 0.4
    ) +
    geom_text(
      aes(label = Cell_label),
      size = 3.5,
      fontface = "bold"
    ) +
    scale_fill_gradient2(
      low = "#2166AC",
      mid = "white",
      high = "#B2182B",
      midpoint = 0,
      limits = c(
        -panel_limit,
        panel_limit
      ),
      labels = label_percent(
        accuracy = 1
      ),
      name = paste0(
        "\u0394 abundance\n",
        "pair − source mean"
      )
    ) +
    labs(
      title = panel_title,
      x = NULL,
      y = NULL
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      plot.title = element_text(
        face = "bold",
        hjust = 0.5,
        size = 10
      ),
      legend.position = "right",
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 7),
      panel.grid = element_blank()
    )
}

dominance_heatmap_panels <- lapply(
  focal_asvs,
  make_dominance_heatmap
)

p_dominant_asv_network <- wrap_plots(
  dominance_heatmap_panels,
  ncol = 4,
  guides = "keep"
) +
  plot_annotation(
    title = paste0(
      "Dominant-ASV abundance changes across ",
      "pairwise coalescence outcomes"
    ),
    subtitle = paste0(
      "W: ASV more abundant in the inferred winner single; ",
      "L: more abundant in the loser single; ",
      "–: no clear dominance"
    )
  )

p_dominant_asv_network

ggsave(
  "Figure_S25C_dominant_ASVs_and_dominance_outcomes.png",
  plot = p_dominant_asv_network,
  width = 14,
  height = 9,
  dpi = 300
)

## ============================================================
## 7. Save results
## ============================================================

write.csv(
  dominance_outcomes,
  "dominance_outcomes_21_pairs.csv",
  row.names = FALSE
)

write.csv(
  dominant_asv_pair_dynamics,
  "Figure_S25C_dominant_ASV_pair_dynamics.csv",
  row.names = FALSE
)

write.csv(
  dominance_asv_concordance_summary,
  "Figure_S25C_dominance_ASV_concordance_summary.csv",
  row.names = FALSE
)

saveRDS(
  dominance_outcomes,
  "dominance_outcomes_21_pairs.rds"
)
```



```{r dominant-asv-neighbourhood-family-and-genus-ranks-2-19, message=FALSE, warning=FALSE}

library(dplyr)
library(tidyr)

## ------------------------------------------------------------
## 1. Add ASV IDs and taxonomy
## ------------------------------------------------------------

asv_long_tax <- asv_long %>%
  left_join(asv_ids, by = "ASV") %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Family, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  mutate(
    Family = ifelse(is.na(Family), "Unassigned", Family),
    Genus = ifelse(is.na(Genus), "Unassigned", Genus)
  )

## ------------------------------------------------------------
## 2. Rank ASVs within each sample
## ------------------------------------------------------------

asv_ranked <- asv_long_tax %>%
  filter(rel_abundance > 0) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()

## ------------------------------------------------------------
## 3. Identify samples where top1 is ASV_1, ASV_2, ASV_9 or ASV_10
## ------------------------------------------------------------

top1_samples <- asv_ranked %>%
  filter(rank == 1) %>%
  select(SampleID, top1_ASV = ASV_ID) %>%
  filter(top1_ASV %in% c("ASV_1", "ASV_2", "ASV_9", "ASV_10"))

## ------------------------------------------------------------
## 4. Function to summarise neighbourhoods for Family or Genus
## ------------------------------------------------------------

make_neighbourhood_table <- function(tax_level) {
  
  tax_level <- rlang::sym(tax_level)
  
  neighbourhood_summary <- asv_ranked %>%
    inner_join(top1_samples, by = "SampleID") %>%
    filter(rank >= 2, rank <= 19) %>%
    group_by(top1_ASV, SampleID, taxon = !!tax_level) %>%
    summarise(
      taxon_abundance_ranks_2_19 = sum(rel_abundance),
      .groups = "drop"
    ) %>%
    group_by(top1_ASV, taxon) %>%
    summarise(
      mean_abundance = mean(taxon_abundance_ranks_2_19),
      .groups = "drop"
    ) %>%
    group_by(top1_ASV) %>%
    arrange(desc(mean_abundance), .by_group = TRUE) %>%
    slice_head(n = 10) %>%
    ungroup()
  
  neighbourhood_table <- neighbourhood_summary %>%
    mutate(
      percent = round(100 * mean_abundance, 2)
    ) %>%
    select(taxon, top1_ASV, percent) %>%
    pivot_wider(
      names_from = top1_ASV,
      values_from = percent
    ) %>%
    arrange(taxon)
  
  list(
    summary = neighbourhood_summary,
    table = neighbourhood_table
  )
}

## ------------------------------------------------------------
## 5. Family neighbourhood table
## ------------------------------------------------------------

family_neighbourhood <- make_neighbourhood_table("Family")

family_neighbourhood_summary <- family_neighbourhood$summary
family_neighbourhood_table <- family_neighbourhood$table

print(family_neighbourhood_table, n = Inf)

write.csv(
  family_neighbourhood_summary,
  "family_neighbourhood_top1_ASV_1_2_9_10_ranks_2_19_summary.csv",
  row.names = FALSE
)

write.csv(
  family_neighbourhood_table,
  "family_neighbourhood_top1_ASV_1_2_9_10_ranks_2_19_table.csv",
  row.names = FALSE
)

## ------------------------------------------------------------
## 6. Genus neighbourhood table
## ------------------------------------------------------------

genus_neighbourhood <- make_neighbourhood_table("Genus")

genus_neighbourhood_summary <- genus_neighbourhood$summary
genus_neighbourhood_table <- genus_neighbourhood$table

print(genus_neighbourhood_table, n = Inf)

write.csv(
  genus_neighbourhood_summary,
  "genus_neighbourhood_top1_ASV_1_2_9_10_ranks_2_19_summary.csv",
  row.names = FALSE
)

write.csv(
  genus_neighbourhood_table,
  "genus_neighbourhood_top1_ASV_1_2_9_10_ranks_2_19_table.csv",
  row.names = FALSE
)

```


```{r pearson-correlation-intra-inter-combinations, message=FALSE, warning=FALSE}


library(phyloseq)
library(dplyr)
library(tibble)
library(ggplot2)
library(cowplot)

## ASV abundance matrix: samples as rows, ASVs as columns
otu_mat <- as(otu_table(ps), "matrix")

if (taxa_are_rows(ps)) {
  otu_mat <- t(otu_mat)
}

## Metadata with explicit SampleID column
metadata_corr <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID") %>%
  select(SampleID, Componentes) %>%
  mutate(
    Complexity = case_when(
      nchar(Componentes) == 1 ~ "Single",
      nchar(Componentes) == 2 ~ "Pair",
      nchar(Componentes) == 3 ~ "Triplet",
      TRUE ~ NA_character_
    ),
    Complexity = factor(Complexity, levels = c("Single", "Pair", "Triplet"))
  )

## Pearson correlation between samples
corr_mat <- cor(t(otu_mat), method = "pearson")

## Convert correlation matrix to pairwise table
sample_pairs <- as.data.frame(as.table(corr_mat)) %>%
  rename(sample1 = Var1, sample2 = Var2, corr = Freq) %>%
  mutate(
    sample1 = as.character(sample1),
    sample2 = as.character(sample2)
  ) %>%
  filter(sample1 < sample2) %>%
  left_join(metadata_corr, by = c("sample1" = "SampleID")) %>%
  rename(
    comb1 = Componentes,
    complexity1 = Complexity
  ) %>%
  left_join(metadata_corr, by = c("sample2" = "SampleID")) %>%
  rename(
    comb2 = Componentes,
    complexity2 = Complexity
  ) %>%
  mutate(
    comparison = ifelse(
      comb1 == comb2,
      "Intra-combination",
      "Inter-combination"
    ),
    intra_complexity = ifelse(
      comparison == "Intra-combination",
      as.character(complexity1),
      NA_character_
    ),
    intra_complexity = factor(
      intra_complexity,
      levels = c("Single", "Pair", "Triplet")
    )
  )

## Summary statistics
corr_stats <- sample_pairs %>%
  group_by(comparison) %>%
  summarise(
    n = n(),
    mean_corr = mean(corr, na.rm = TRUE),
    sd_corr = sd(corr, na.rm = TRUE),
    median_corr = median(corr, na.rm = TRUE),
    q25_corr = quantile(corr, 0.25, na.rm = TRUE),
    q75_corr = quantile(corr, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

print(corr_stats)

## Main density plot
p_corr_density <- ggplot(sample_pairs,
                         aes(x = corr,
                             fill = comparison,
                             color = comparison)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  scale_fill_manual(values = c(
    "Intra-combination" = "#4D4D4D",
    "Inter-combination" = "#BDBDBD"
  )) +
  scale_color_manual(values = c(
    "Intra-combination" = "#252525",
    "Inter-combination" = "#737373"
  )) +
  theme_bw() +
  labs(
    x = "Pearson correlation",
    y = "Density",
    title = "Within- vs between-combination Pearson correlations",
    fill = "Comparison",
    color = "Comparison"
  )

## Inset
p_inset <- sample_pairs %>%
  filter(comparison == "Intra-combination") %>%
  ggplot(aes(x = corr,
             fill = intra_complexity,
             color = intra_complexity)) +
  geom_density(alpha = 0.30, linewidth = 0.6) +
  scale_fill_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  scale_color_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  theme_bw(base_size = 8) +
  labs(
    x = NULL,
    y = NULL,
    title = "Intra only",
    fill = "Complexity",
    color = "Complexity"
  ) +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    plot.title = element_text(size = 9),
    axis.text = element_text(size = 7),
    panel.grid.minor = element_blank()
  )

## Combined plot
p_corr_density_inset <- ggdraw() +
  draw_plot(p_corr_density) +
  draw_plot(
    p_inset,
    x = 0.09,
    y = 0.55,
    width = 0.38,
    height = 0.38
  )

p_corr_density_inset

ggsave(
  "pearson_correlation_intra_inter_density_with_inset.png",
  plot = p_corr_density_inset,
  width = 8,
  height = 6,
  dpi = 300
)




## ------------------------------------------------------------
## 7. Final version: no title + legend inside below inset
## ------------------------------------------------------------

p_corr_density_final <- ggplot(sample_pairs,
                               aes(x = corr,
                                   fill = comparison,
                                   color = comparison)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  scale_fill_manual(values = c(
    "Intra-combination" = "#4D4D4D",
    "Inter-combination" = "#BDBDBD"
  )) +
  scale_color_manual(values = c(
    "Intra-combination" = "#252525",
    "Inter-combination" = "#737373"
  )) +
  theme_bw() +
  labs(
    x = "Pearson correlation",
    y = "Density",
    title = NULL,
    fill = "Comparison",
    color = "Comparison"
  ) +
  theme(
    legend.position = c(0.28, 0.4),
    legend.background = element_rect(
      fill = scales::alpha("white", 0.75),
      colour = NA
    ),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

p_corr_density_inset_final <- ggdraw() +
  draw_plot(p_corr_density_final) +
  draw_plot(
    p_inset,
    x = 0.09,
    y = 0.55,
    width = 0.38,
    height = 0.38
  )

p_corr_density_inset_final

ggsave(
  "pearson_correlation_intra_inter_density_with_inset_final.png",
  plot = p_corr_density_inset_final,
  width = 8,
  height = 6,
  dpi = 300
)

```


 ### Distance to the compositional Bray–Curtis centroid.
 

```{r balanced-loto-centroid}
library(phyloseq)
library(dplyr)
library(tibble)
library(ggplot2)
library(vegan)

## ------------------------------------------------------------
## 1. Relative-abundance matrix
## ------------------------------------------------------------

otu_mat <- as(otu_table(ps), "matrix")

if (taxa_are_rows(ps)) {
  otu_mat <- t(otu_mat)
}

## Remove samples with no reads, if any
otu_mat <- otu_mat[rowSums(otu_mat) > 0, , drop = FALSE]

## Convert each sample to relative abundances
otu_rel <- sweep(
  otu_mat,
  1,
  rowSums(otu_mat),
  "/"
)

## ------------------------------------------------------------
## 2. Metadata and complexity class
## ------------------------------------------------------------

metadata_loto <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID") %>%
  mutate(
    Treatment = as.character(Componentes),
    Complexity = case_when(
      nchar(Treatment) == 1 ~ "Single",
      nchar(Treatment) == 2 ~ "Pair",
      nchar(Treatment) == 3 ~ "Triplet",
      TRUE ~ NA_character_
    ),
    Complexity = factor(
      Complexity,
      levels = c("Single", "Pair", "Triplet")
    )
  ) %>%
  filter(
    !is.na(Complexity),
    SampleID %in% rownames(otu_rel)
  )

## Arrange metadata in the same order as the abundance matrix
metadata_loto <- metadata_loto[
  match(rownames(otu_rel), metadata_loto$SampleID),
  ,
  drop = FALSE
]

stopifnot(
  identical(
    rownames(otu_rel),
    metadata_loto$SampleID
  )
)

## ------------------------------------------------------------
## 3. Mean composition of each treatment combination
## ------------------------------------------------------------

treatment_ids <- unique(metadata_loto$Treatment)

treatment_mean_composition <- do.call(
  rbind,
  lapply(treatment_ids, function(current_treatment) {

    treatment_samples <- metadata_loto$SampleID[
      metadata_loto$Treatment == current_treatment
    ]

    colMeans(
      otu_rel[treatment_samples, , drop = FALSE],
      na.rm = TRUE
    )
  })
)

rownames(treatment_mean_composition) <- treatment_ids

## One metadata row per treatment combination
treatment_metadata <- metadata_loto %>%
  distinct(Treatment, Complexity) %>%
  arrange(
    match(
      Treatment,
      rownames(treatment_mean_composition)
    )
  )

stopifnot(
  identical(
    treatment_metadata$Treatment,
    rownames(treatment_mean_composition)
  )
)

## Check the expected number of treatment combinations:
## 7 singles, 21 pairs and 35 triplets
print(table(treatment_metadata$Complexity))

## ------------------------------------------------------------
## 4. Balanced leave-one-treatment-out centroid
## ------------------------------------------------------------

calculate_loto_distance <- function(treatment_to_test) {

  tested_class <- as.character(
    treatment_metadata$Complexity[
      treatment_metadata$Treatment == treatment_to_test
    ]
  )

  ## Calculate one centroid for each complexity class.
  ## The tested treatment is excluded from its own class.
  class_centroids <- lapply(
    levels(treatment_metadata$Complexity),
    function(current_class) {

      class_treatments <- treatment_metadata$Treatment[
        treatment_metadata$Complexity == current_class
      ]

      if (current_class == tested_class) {
        class_treatments <- setdiff(
          class_treatments,
          treatment_to_test
        )
      }

      if (length(class_treatments) == 0) {
        stop(
          paste(
            "No treatments remain in class",
            current_class,
            "after leave-one-treatment-out exclusion."
          )
        )
      }

      ## Equal weight for every treatment within the class
      colMeans(
        treatment_mean_composition[
          class_treatments,
          ,
          drop = FALSE
        ],
        na.rm = TRUE
      )
    }
  )

  ## Equal weight for Single, Pair and Triplet:
  ## each class contributes one third to the reference centroid
  balanced_loto_centroid <- Reduce(
    "+",
    class_centroids
  ) / length(class_centroids)

  ## Bray-Curtis distance between the held-out treatment mean
  ## and its independently calculated balanced centroid
  as.numeric(
    vegan::vegdist(
      rbind(
        Held_out_treatment =
          treatment_mean_composition[
            treatment_to_test,
            ,
            drop = FALSE
          ],
        Balanced_LOTO_centroid =
          balanced_loto_centroid
      ),
      method = "bray"
    )
  )
}

treatment_metadata$distance_to_balanced_loto_centroid <- vapply(
  treatment_metadata$Treatment,
  calculate_loto_distance,
  numeric(1)
)

## ------------------------------------------------------------
## 5. Inspect and save treatment-level results
## ------------------------------------------------------------

print(treatment_metadata)

summary_by_complexity <- treatment_metadata %>%
  group_by(Complexity) %>%
  summarise(
    n_treatments = n(),
    mean_distance =
      mean(distance_to_balanced_loto_centroid),
    median_distance =
      median(distance_to_balanced_loto_centroid),
    sd_distance =
      sd(distance_to_balanced_loto_centroid),
    .groups = "drop"
  )

print(summary_by_complexity)

write.csv(
  treatment_metadata,
  "balanced_LOTO_centroid_distances_by_treatment.csv",
  row.names = FALSE
)

write.csv(
  summary_by_complexity,
  "balanced_LOTO_centroid_summary_by_complexity.csv",
  row.names = FALSE
)

save(
  treatment_metadata,
  treatment_mean_composition,
  summary_by_complexity,
  file = "balanced_LOTO_centroid_analysis.RData"
)

## ------------------------------------------------------------
## 6. Density plot for all three complexity classes
## ------------------------------------------------------------

complexity_colors <- c(
  "Single"  = "#1b9e77",
  "Pair"    = "#d95f02",
  "Triplet" = "#7570b3"
)

p_balanced_loto_centroid <- ggplot(
  treatment_metadata,
  aes(
    x = distance_to_balanced_loto_centroid,
    fill = Complexity,
    color = Complexity
  )
) +
  geom_density(
    alpha = 0.25,
    linewidth = 0.8,
    adjust = 1
  ) +
  scale_color_manual(
    values = complexity_colors,
    drop = FALSE
  ) +
  scale_fill_manual(
    values = complexity_colors,
    drop = FALSE
  ) +
  labs(
    x = paste0(
      "Bray-Curtis distance to balanced ",
      "leave-one-treatment-out centroid"
    ),
    y = "Density",
    color = "Complexity",
    fill = "Complexity"
  ) +
  theme_bw() +
  theme(
    legend.position = c(0.78, 0.80),
    legend.background = element_rect(
      fill = scales::alpha("white", 0.75),
      colour = NA
    )
  )

p_balanced_loto_centroid

ggsave(
  "bray_curtis_distance_to_balanced_LOTO_centroid.png",
  plot = p_balanced_loto_centroid,
  width = 8,
  height = 6,
  dpi = 300
)
```
```
 




### ASV-level assembly trait space in 2D

This chunk uses the filtered ASV table, retaining only ASVs reaching at least 0.1% relative abundance in at least two samples. For each retained ASV, the source set is defined from the single-inoculum communities. We then calculate three assembly axes: conditional persistence, mean retention in eligible mixtures, and context dependence. The resulting ASV-level trait space is visualized as three 2D plots. Each ASV is shown by its numeric ASV label, with text size proportional to source breadth and colour indicating taxonomic family.
```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

## ------------------------------------------------------------
## 1. Filter ASVs: >=0.1% relative abundance in at least 2 samples
## ------------------------------------------------------------

asv_filter_summary <- asv_long %>%
  group_by(ASV) %>%
  summarise(
    n_samples_above_threshold = sum(rel_abundance >= 0.001),
    .groups = "drop"
  )

asvs_kept <- asv_filter_summary %>%
  filter(n_samples_above_threshold >= 2) %>%
  pull(ASV)

asv_long_filt <- asv_long %>%
  filter(ASV %in% asvs_kept)

## ------------------------------------------------------------
## 2. Add ASV_ID and taxonomy
## ------------------------------------------------------------

taxonomy_info <- asv_context_summary_tax %>%
  select(ASV, Family, Genus) %>%
  distinct() %>%
  mutate(
    Family = ifelse(is.na(Family), "Unassigned", Family),
    Genus = ifelse(is.na(Genus), "Unassigned", Genus)
  )

asv_long_filt <- asv_long_filt %>%
  left_join(asv_ids, by = "ASV") %>%
  left_join(taxonomy_info, by = "ASV") %>%
  mutate(
    ASV_number = as.numeric(gsub("ASV_", "", ASV_ID))
  )

## ------------------------------------------------------------
## 3. Mean abundance per ASV x context
## ------------------------------------------------------------

asv_context_mean <- asv_long_filt %>%
  group_by(ASV, ASV_ID, ASV_number, Family, Genus, Componentes) %>%
  summarise(
    mean_abundance = mean(rel_abundance),
    present = mean_abundance >= 0.001,
    .groups = "drop"
  ) %>%
  mutate(
    context_size = nchar(Componentes)
  )

## ------------------------------------------------------------
## 4. Define source set from singles
## ------------------------------------------------------------

source_set <- asv_context_mean %>%
  filter(context_size == 1, present) %>%
  group_by(ASV, ASV_ID, ASV_number, Family, Genus) %>%
  summarise(
    source_components = list(Componentes),
    source_breadth = n(),
    .groups = "drop"
  )

## Keep only ASVs detected in at least one single inoculum
asv_context_mean_sources <- asv_context_mean %>%
  inner_join(
    source_set,
    by = c("ASV", "ASV_ID", "ASV_number", "Family", "Genus")
  )

## ------------------------------------------------------------
## 5. Evaluate only eligible mixtures
## ------------------------------------------------------------

eligible_mixtures <- asv_context_mean_sources %>%
  filter(context_size %in% c(2, 3)) %>%
  rowwise() %>%
  mutate(
    mixture_components = list(strsplit(Componentes, "")[[1]]),
    eligible = length(
      intersect(
        unlist(source_components),
        unlist(mixture_components)
      )
    ) > 0
  ) %>%
  ungroup() %>%
  filter(eligible)

## ------------------------------------------------------------
## 6. Add best source-single abundance for each eligible mixture
## ------------------------------------------------------------

single_abundances <- asv_context_mean_sources %>%
  filter(context_size == 1) %>%
  select(
    ASV,
    single_component = Componentes,
    single_abundance = mean_abundance
  )

eligible_mixtures <- eligible_mixtures %>%
  rowwise() %>%
  mutate(
    best_source_single_abundance = max(
      single_abundances$single_abundance[
        single_abundances$ASV == ASV &
          single_abundances$single_component %in% unlist(mixture_components)
      ],
      na.rm = TRUE
    )
  ) %>%
  ungroup()

## ------------------------------------------------------------
## 7. Calculate retention per ASV x eligible mixture
## ------------------------------------------------------------

epsilon <- 1e-6

eligible_mixtures <- eligible_mixtures %>%
  mutate(
    retention = log2(
      (mean_abundance + epsilon) /
        (best_source_single_abundance + epsilon)
    )
  )

## ------------------------------------------------------------
## 8. ASV-level assembly axes
## ------------------------------------------------------------

asv_trait_space <- eligible_mixtures %>%
  group_by(ASV, ASV_ID, ASV_number, Family, Genus, source_breadth) %>%
  summarise(
    n_eligible_mixtures = n(),
    conditional_persistence = mean(present),
    mean_retention = mean(retention, na.rm = TRUE),
    context_dependence = sd(retention, na.rm = TRUE),
    max_abundance_in_mixtures = max(mean_abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    context_dependence = ifelse(is.na(context_dependence), 0, context_dependence)
  )

print(asv_trait_space, n = 30)

write.csv(
  asv_trait_space,
  "asv_assembly_trait_space.csv",
  row.names = FALSE
)

## ------------------------------------------------------------
## 9. Collapse taxonomy to top 10 families + Other
## ------------------------------------------------------------

top10_families <- asv_trait_space %>%
  group_by(Family) %>%
  summarise(
    total_weight = sum(max_abundance_in_mixtures, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_weight)) %>%
  slice_head(n = 10) %>%
  pull(Family)

asv_trait_space_plot <- asv_trait_space %>%
  mutate(
    Family_plot = ifelse(Family %in% top10_families, Family, "Other")
  )

## palette
family_levels <- c(sort(setdiff(unique(asv_trait_space_plot$Family_plot), "Other")), "Other")

family_colors <- setNames(
  c(scales::hue_pal()(length(family_levels)-1), "grey75"),
  family_levels
)

## ------------------------------------------------------------
## 10. Plot 1: retention vs persistence
## ------------------------------------------------------------

p1 <- ggplot(
  asv_trait_space_plot,
  aes(
    x = mean_retention,
    y = conditional_persistence,
    size = source_breadth,
    color = Family_plot
  )
) +
  geom_point(alpha = 0.75) +
  scale_color_manual(values = family_colors) +
  theme_bw() +
  labs(
    x = "Mean retention",
    y = "Conditional persistence",
    #title = "Retention vs persistence",
    color = "Family",
    size = "Source breadth"
  )

## ------------------------------------------------------------
## 11. Plot 2: retention vs context dependence
## ------------------------------------------------------------

p2 <- ggplot(
  asv_trait_space_plot,
  aes(
    x = mean_retention,
    y = context_dependence,
    size = source_breadth,
    color = Family_plot
  )
) +
  geom_point(alpha = 0.75) +
  scale_color_manual(values = family_colors) +
  theme_bw() +
  labs(
    x = "Mean retention",
    y = "Context dependence",
    #title = "Retention vs context dependence",
    color = "Family",
    size = "Source breadth"
  )

## ------------------------------------------------------------
## 12. Combine with shared legend
## ------------------------------------------------------------

p_combined <- (p1 + p2) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p_combined

ggsave(
  "asv_trait_space_2plots.png",
  plot = p_combined,
  width = 12,
  height = 6,
  dpi = 300
)

```






