---
title: "New analyses"
author: "Daniel Aguirre"
date: "2026-06-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(phyloseq)
library(dplyr)
library(tibble)

setwd("C:/Users/DA.5015582/OneDrive - UAM/TRABAJO/Abiertos/Learning/Assembly")

ps <- readRDS("ps_nocontrol_filtrado_rare.rds")
metadata <- data.frame(sample_data(ps))
otu_mat <- as(otu_table(ps), "matrix")
if (!taxa_are_rows(ps)) otu_mat <- t(otu_mat)
taxonomy <- as.data.frame(tax_table(ps))

asv_prevalence <- rowSums(otu_mat > 0)
asv_total_abundance <- rowSums(otu_mat)
```

```{r create-asv-context-table, message=FALSE, warning=FALSE}

library(tidyr)

## Extract metadata
metadata <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID") %>%
  select(SampleID, Componentes)

## Extract ASV table
otu_mat <- as(otu_table(ps), "matrix")

if (!taxa_are_rows(ps)) {
  otu_mat <- t(otu_mat)
}

## Convert counts to relative abundance
## Since all samples are rarefied to the same depth, this is counts / 23394
otu_rel <- sweep(otu_mat, 2, colSums(otu_mat), "/")

## Long ASV table: one row per ASV x sample
asv_long <- as.data.frame(otu_rel) %>%
  rownames_to_column("ASV") %>%
  pivot_longer(
    cols = -ASV,
    names_to = "SampleID",
    values_to = "rel_abundance"
  ) %>%
  left_join(metadata, by = "SampleID")

## Summary per ASV x inoculum context
asv_context_summary <- asv_long %>%
  group_by(ASV, Componentes) %>%
  summarise(
    n_samples = n(),
    prevalence = sum(rel_abundance > 0),
    prevalence_prop = prevalence / n_samples,
    mean_abundance = mean(rel_abundance),
    median_abundance = median(rel_abundance),
    max_abundance = max(rel_abundance),
    .groups = "drop"
  )

## Add taxonomy
taxonomy <- as.data.frame(tax_table(ps)) %>%
  rownames_to_column("ASV")

asv_context_summary_tax <- asv_context_summary %>%
  left_join(taxonomy, by = "ASV")

## Quick checks
dim(asv_long)
dim(asv_context_summary_tax)

print(head(asv_long, 10))
print(head(asv_context_summary_tax, 10))

metadata_check <- data.frame(sample_data(ps)) |>
  tibble::rownames_to_column("SampleID")

table(metadata_check$Componentes)

asv_context_summary_tax |>
  dplyr::filter(mean_abundance > 0) |>
  dplyr::arrange(desc(mean_abundance)) |>
  print(n = 20)

## Save tables
write.csv(asv_long, "asv_long_relative_abundance.csv", row.names = FALSE)
write.csv(asv_context_summary_tax, "asv_context_summary_tax.csv", row.names = FALSE)

```

```{r}

## Filter ASVs already present in asv_long / asv_context_summary_tax
## Criterion: retain ASVs reaching >= 0.1% relative abundance in at least 2 samples.
## Rationale: this removes very rare/spurious ASVs while keeping taxa with evidence of reproducible ecological establishment.

#RESULT= 423 ASVs
asv_filter_summary <- asv_long %>%
  group_by(ASV) %>%
  summarise(
    n_samples_above_0.1percent = sum(rel_abundance >= 0.001),
    max_rel_abundance = max(rel_abundance),
    .groups = "drop"
  )

asvs_kept <- asv_filter_summary %>%
  filter(n_samples_above_0.1percent >= 2) %>%
  pull(ASV)

asv_long_filtered <- asv_long %>%
  filter(ASV %in% asvs_kept)

asv_context_summary_tax_filtered <- asv_context_summary_tax %>%
  filter(ASV %in% asvs_kept)

cat("Original ASVs:", length(unique(asv_long$ASV)), "\n")
cat("Filtered ASVs:", length(unique(asv_long_filtered$ASV)), "\n")

write.csv(asv_long_filtered,
          "asv_long_relative_abundance_filtered.csv",
          row.names = FALSE)

write.csv(asv_context_summary_tax_filtered,
          "asv_context_summary_tax_filtered.csv",
          row.names = FALSE)

```

```{r rank-abundance-global}
## ============================================================
## Rank-abundance analyses using the unfiltered ASV table
## Input object: asv_long
## ============================================================

library(dplyr)
library(ggplot2)
library(cowplot)

## ------------------------------------------------------------
## 1. Global rank-abundance curves, all samples together
## ------------------------------------------------------------

rank_abundance <- asv_long %>%
  filter(rel_abundance > 0) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()

p_rank_all <- ggplot(rank_abundance,
                     aes(x = rank,
                         y = rel_abundance,
                         group = SampleID)) +
  geom_line(alpha = 0.05, linewidth = 0.3) +
  scale_y_log10() +
  labs(
    x = "Rank",
    y = "Relative abundance",
    title = "Global rank-abundance distribution"
  ) +
  theme_bw()

p_rank_all

ggsave(
  "rank_abundance_all_unfiltered.png",
  plot = p_rank_all,
  width = 8,
  height = 6,
  dpi = 300
)


## ------------------------------------------------------------
## 2. Global rank-abundance envelope, all samples together
## ------------------------------------------------------------

rank_summary <- rank_abundance %>%
  group_by(rank) %>%
  summarise(
    q05 = quantile(rel_abundance, 0.05, na.rm = TRUE),
    q25 = quantile(rel_abundance, 0.25, na.rm = TRUE),
    median = median(rel_abundance, na.rm = TRUE),
    q75 = quantile(rel_abundance, 0.75, na.rm = TRUE),
    q95 = quantile(rel_abundance, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

p_envelope <- ggplot(rank_summary, aes(x = rank)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.15) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.3) +
  geom_line(aes(y = median), linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Rank",
    y = "Relative abundance",
    title = "Rank-abundance envelope across all samples"
  ) +
  theme_bw()

p_envelope

ggsave(
  "rank_abundance_all_envelope_unfiltered.png",
  plot = p_envelope,
  width = 8,
  height = 6,
  dpi = 300
)


## ------------------------------------------------------------
## 3. Rank-abundance envelopes by inoculum complexity
## ------------------------------------------------------------

rank_abundance_complexity <- asv_long %>%
  filter(rel_abundance > 0) %>%
  mutate(
    Complexity = case_when(
      nchar(Componentes) == 1 ~ "Single",
      nchar(Componentes) == 2 ~ "Pair",
      nchar(Componentes) == 3 ~ "Triplet",
      TRUE ~ NA_character_
    ),
    Complexity = factor(Complexity, levels = c("Single", "Pair", "Triplet"))
  ) %>%
  filter(!is.na(Complexity)) %>%
  group_by(SampleID, Complexity) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()

rank_summary_complexity <- rank_abundance_complexity %>%
  group_by(Complexity, rank) %>%
  summarise(
    q05 = quantile(rel_abundance, 0.05, na.rm = TRUE),
    q25 = quantile(rel_abundance, 0.25, na.rm = TRUE),
    median = median(rel_abundance, na.rm = TRUE),
    q75 = quantile(rel_abundance, 0.75, na.rm = TRUE),
    q95 = quantile(rel_abundance, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

p_envelope_complexity <- ggplot(rank_summary_complexity, aes(x = rank)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.15) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.3) +
  geom_line(aes(y = median), linewidth = 0.8) +
  scale_y_log10() +
  facet_grid(Complexity ~ ., scales = "fixed") +
  labs(
    x = "Rank",
    y = "Relative abundance",
    title = "Rank-abundance envelopes by inoculum complexity"
  ) +
  theme_bw()


p_envelope_complexity

ggsave(
  "rank_abundance_envelope_by_complexity_unfiltered.png",
  plot = p_envelope_complexity,
  width = 8,
  height = 10,
  dpi = 300
)


## ------------------------------------------------------------
## 4. Overlapping envelopes by inoculum complexity
## ------------------------------------------------------------
# 
# p_envelope_overlap <- ggplot(rank_summary_complexity,
#                              aes(x = rank,
#                                  color = Complexity,
#                                  fill = Complexity)) +
#   geom_ribbon(aes(ymin = q05, ymax = q95),
#               alpha = 0.12,
#               color = NA) +
#   geom_ribbon(aes(ymin = q25, ymax = q75),
#               alpha = 0.22,
#               color = NA) +
#   geom_line(aes(y = median),
#             linewidth = 0.9) +
#   scale_y_log10() +
#   scale_color_manual(values = c(
#     "Single" = "#1b9e77",
#     "Pair" = "#d95f02",
#     "Triplet" = "#7570b3"
#   )) +
#   scale_fill_manual(values = c(
#     "Single" = "#1b9e77",
#     "Pair" = "#d95f02",
#     "Triplet" = "#7570b3"
#   )) +
#   labs(
#     x = "Rank",
#     y = "Relative abundance",
#     title = "Overlapping rank-abundance envelopes by inoculum complexity",
#     color = "Complexity",
#     fill = "Complexity"
#   ) +
#   theme_bw()
# 
# p_envelope_overlap
# 
# ggsave(
#   "rank_abundance_envelope_complexity_overlap_unfiltered.png",
#   plot = p_envelope_overlap,
#   width = 8,
#   height = 6,
#   dpi = 300
# )




p_envelope_overlap <- ggplot(rank_summary_complexity,
                             aes(x = rank,
                                 color = Complexity,
                                 fill = Complexity)) +
  geom_ribbon(aes(ymin = q05, ymax = q95),
              alpha = 0.12,
              color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75),
              alpha = 0.22,
              color = NA) +
  geom_line(aes(y = median),
            linewidth = 0.9) +
  scale_y_log10() +
  scale_color_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  scale_fill_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  labs(
    x = "Rank",
    y = "Relative abundance",
    title = NULL,
    color = "Complexity",
    fill = "Complexity"
  ) +
  theme_bw() +
  theme(
    legend.position = c(0.66, 0.35),
    legend.background = element_rect(
      fill = scales::alpha("white", 0.75),
      colour = NA
    )
  )

## ------------------------------------------------------------
## 5. Inset plot: same analysis but only ASVs >1%
## ------------------------------------------------------------

rank_abundance_gt1 <- asv_long %>%
  filter(rel_abundance > 0.01) %>%
  mutate(
    Complexity = case_when(
      nchar(Componentes) == 1 ~ "Single",
      nchar(Componentes) == 2 ~ "Pair",
      nchar(Componentes) == 3 ~ "Triplet",
      TRUE ~ NA_character_
    ),
    Complexity = factor(Complexity, levels = c("Single", "Pair", "Triplet"))
  ) %>%
  filter(!is.na(Complexity)) %>%
  group_by(SampleID, Complexity) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()

rank_summary_gt1 <- rank_abundance_gt1 %>%
  group_by(Complexity, rank) %>%
  summarise(
    q05 = quantile(rel_abundance, 0.05, na.rm = TRUE),
    q25 = quantile(rel_abundance, 0.25, na.rm = TRUE),
    median = median(rel_abundance, na.rm = TRUE),
    q75 = quantile(rel_abundance, 0.75, na.rm = TRUE),
    q95 = quantile(rel_abundance, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

p_inset <- ggplot(rank_summary_gt1,
                  aes(x = rank,
                      color = Complexity,
                      fill = Complexity)) +
  geom_ribbon(aes(ymin = q05, ymax = q95),
              alpha = 0.12,
              color = NA) +
  geom_ribbon(aes(ymin = q25, ymax = q75),
              alpha = 0.22,
              color = NA) +
  geom_line(aes(y = median), linewidth = 0.7) +
  scale_y_log10() +
  scale_color_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  scale_fill_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  labs(
    x = "Rank",
    y = NULL,
    title = ">1%"
  ) +
  theme_bw(base_size = 8) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 9),
    axis.title.x = element_text(size = 8),
    axis.text = element_text(size = 7)
  )

p_combined <- ggdraw() +
  draw_plot(p_envelope_overlap) +
  draw_plot(
    p_inset,
    x = 0.45,
    y = 0.52,
    width = 0.40,
    height = 0.40
  )

p_combined

ggsave(
  "rank_abundance_envelope_complexity_overlap_with_gt1_inset_unfiltered.png",
  plot = p_combined,
  width = 8,
  height = 6,
  dpi = 300
)








## ------------------------------------------------------------
## 6. Cumulative abundance: ASVs needed to reach 50% and 90%
## ------------------------------------------------------------

cumulative_rank_abundance <- asv_long %>%
  filter(rel_abundance > 0) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  mutate(
    rank = row_number(),
    cumulative_abundance = cumsum(rel_abundance)
  ) %>%
  ungroup()

n_asvs_summary <- cumulative_rank_abundance %>%
  group_by(SampleID) %>%
  summarise(
    n_asvs_50 = min(rank[cumulative_abundance >= 0.50]),
    n_asvs_90 = min(rank[cumulative_abundance >= 0.90]),
    .groups = "drop"
  )

summary_stats <- n_asvs_summary %>%
  summarise(
    median_50 = median(n_asvs_50, na.rm = TRUE),
    q25_50 = quantile(n_asvs_50, 0.25, na.rm = TRUE),
    q75_50 = quantile(n_asvs_50, 0.75, na.rm = TRUE),
    median_90 = median(n_asvs_90, na.rm = TRUE),
    q25_90 = quantile(n_asvs_90, 0.25, na.rm = TRUE),
    q75_90 = quantile(n_asvs_90, 0.75, na.rm = TRUE)
  )

print(summary_stats)
```

```{r bray-nmds-pcoa, message=FALSE, warning=FALSE}

set.seed(123)

## ============================================================
## 1. NMDS with all samples
##    Bray-Curtis, labels only, no points
## ============================================================

ord_nmds_bray_all <- ordinate(
  ps,
  method = "NMDS",
  distance = "bray",
  trymax = 100
)


## ============================================================
## 2. NMDS using only single-inoculum samples
##    Important: subset first, then run NMDS on this subset
## ============================================================

ps_singles <- subset_samples(ps, nchar(Componentes) == 1)

## Optional: remove ASVs absent from the singles subset
ps_singles <- prune_taxa(taxa_sums(ps_singles) > 0, ps_singles)

ord_nmds_bray_singles <- ordinate(
  ps_singles,
  method = "NMDS",
  distance = "bray",
  trymax = 100
)



ord_df_all <- plot_ordination(ps, ord_nmds_bray_all, type = "samples", justDF = TRUE)

p_nmds_bray_all <- ggplot(ord_df_all, aes(x = NMDS1, y = NMDS2)) +
  geom_text(aes(label = Componentes), size = 3, check_overlap = TRUE) +
  labs(title = "NMDS - Bray-Curtis, all samples") +
  theme_bw()

ggsave(
  "NMDS_Bray_Curtis_all_labels_only.png",
  plot = p_nmds_bray_all,
  width = 8,
  height = 6,
  dpi = 300
)

ord_df_singles <- plot_ordination(ps_singles, ord_nmds_bray_singles,
                                  type = "samples", justDF = TRUE)

p_nmds_bray_singles <- ggplot(ord_df_singles,
                              aes(x = NMDS1, y = NMDS2)) +
  geom_text(aes(label = Componentes, color = Componentes),
            size = 4, check_overlap = TRUE) +
  labs(title = "NMDS - Bray-Curtis, single inocula only",
       color = "Component") +
  theme_bw()

ggsave(
  "NMDS_Bray_Curtis_singles_labels_only.png",
  plot = p_nmds_bray_singles,
  width = 8,
  height = 6,
  dpi = 300
)


pA <- p_nmds_bray_singles +
  labs(title = NULL, color = NULL) +
  theme(legend.position = "none") +
  annotate("text", x = -Inf, y = Inf, label = "A",
           hjust = -0.7, vjust = 1.5, fontface = "bold", size = 6)

pB <- p_nmds_bray_all +
  labs(title = NULL) +
  theme(legend.position = "none") +
  annotate("text", x = -Inf, y = Inf, label = "B",
           hjust = -0.7, vjust = 1.5, fontface = "bold", size = 6)

p_nmds_combined <- cowplot::plot_grid(
  pA, pB,
  nrow = 1,
  labels = NULL,
  align = "hv",
  rel_widths = c(1, 1.25)
)

ggsave(
  "NMDS_Bray_Curtis_singles_and_all.png",
  plot = p_nmds_combined,
  width = 12,
  height = 5,
  dpi = 300
)

```

```{r, message=FALSE, warning=FALSE}
top2_pair_freq <- asv_long %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  filter(rel_abundance > 0) %>%
  mutate(Genus = ifelse(is.na(Genus), "Unassigned", Genus)) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  select(SampleID, rank, Genus) %>%
  tidyr::pivot_wider(
    names_from = rank,
    values_from = Genus,
    names_prefix = "rank"
  ) %>%
  mutate(pair = paste(rank1, rank2, sep = " - ")) %>%
  count(pair, name = "n_samples") %>%
  mutate(percent = 100 * n_samples / sum(n_samples)) %>%
  arrange(desc(percent))

print(top2_pair_freq, n = 20)

#No Pseudomonas domination:

non_pseudomonas_top2 <- asv_long %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  filter(rel_abundance > 0) %>%
  mutate(Genus = ifelse(is.na(Genus), "Unassigned", Genus)) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  summarise(
    top1 = Genus[1],
    top2 = Genus[2],
    Componentes = first(Componentes),
    both_non_pseudomonas = all(Genus != "Pseudomonas"),
    .groups = "drop"
  ) %>%
  filter(both_non_pseudomonas)

print(non_pseudomonas_top2, n = Inf)

top2_ACD <- asv_long %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  filter(rel_abundance > 0) %>%
  filter(Componentes %in% c("A", "C", "D")) %>%
  mutate(Genus = ifelse(is.na(Genus), "Unassigned", Genus)) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  mutate(rank = row_number()) %>%
  ungroup()

print(top2_ACD, n = Inf)



asv_ids <- asv_long %>%
  group_by(ASV) %>%
  summarise(global_abundance = sum(rel_abundance), .groups = "drop") %>%
  arrange(desc(global_abundance)) %>%
  mutate(ASV_ID = paste0("ASV_", row_number()))

print(head(asv_ids, 20))



top2_asvs_global <- asv_long %>%
  left_join(asv_ids, by = "ASV") %>%
  filter(rel_abundance > 0) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  count(rank, ASV_ID, name = "n_samples") %>%
  group_by(rank) %>%
  mutate(percent = 100 * n_samples / sum(n_samples)) %>%
  arrange(rank, desc(percent))

print(top2_asvs_global, n = 20)


top2_selected <- asv_long %>%
  left_join(asv_ids, by = "ASV") %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  filter(rel_abundance > 0) %>%
  filter(Componentes %in% c("G", "A", "C", "D", "CG", "ADG")) %>%
  mutate(Genus = ifelse(is.na(Genus), "Unassigned", Genus)) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>%
  select(SampleID, Componentes, rank, ASV_ID, Genus, rel_abundance)

top2_selected_wide <- top2_selected %>%
  tidyr::pivot_wider(
    names_from = rank,
    values_from = c(ASV_ID, Genus, rel_abundance),
    names_glue = "{.value}_rank{rank}"
  ) %>%
  arrange(Componentes)

print(top2_selected_wide, n = Inf)


asv_focus_singles <- asv_long %>%
  left_join(asv_ids, by = "ASV") %>%
  filter(nchar(Componentes) == 1) %>%
  filter(ASV_ID %in% c("ASV_1", "ASV_2", "ASV_9", "ASV_10")) %>%
  select(SampleID, Componentes, ASV_ID, rel_abundance)

print(asv_focus_singles, n = Inf)

ggplot(asv_focus_singles,
       aes(x = Componentes,
           y = rel_abundance,
           fill = ASV_ID)) +
  geom_boxplot() +
  facet_wrap(~ASV_ID, scales = "free_y") +
  theme_bw()


pseudomonas_abundance_top10 <- asv_long %>%
  group_by(ASV) %>%
  summarise(mean_rel_abundance = mean(rel_abundance), .groups = "drop") %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  filter(Genus == "Pseudomonas") %>%
  left_join(asv_ids, by = "ASV") %>%
  arrange(as.numeric(gsub("ASV_", "", ASV_ID))) %>%
  slice_head(n = 10)

p_pseudomonas_top10 <- ggplot(
  pseudomonas_abundance_top10,
  aes(
    x = reorder(ASV_ID, as.numeric(gsub("ASV_", "", ASV_ID))),
    y = mean_rel_abundance
  )
) +
  geom_col() +
  geom_text(
    aes(label = round(mean_rel_abundance, 3)),
    vjust = -0.3,
    size = 3
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    x = "Pseudomonas ASV",
    y = "Mean relative abundance",
    title = "Top 10 Pseudomonas ASVs (out of 186 total)"
  )

p_pseudomonas_top10

ggsave(
  "top10_pseudomonas_asvs_mean_relative_abundance.png",
  plot = p_pseudomonas_top10,
  width = 8,
  height = 5,
  dpi = 300
)
```

```{r dominant-asv-neighbourhood-family-and-genus-ranks-2-19, message=FALSE, warning=FALSE}

library(dplyr)
library(tidyr)

## ------------------------------------------------------------
## 1. Add ASV IDs and taxonomy
## ------------------------------------------------------------

asv_long_tax <- asv_long %>%
  left_join(asv_ids, by = "ASV") %>%
  left_join(
    asv_context_summary_tax %>%
      select(ASV, Family, Genus) %>%
      distinct(),
    by = "ASV"
  ) %>%
  mutate(
    Family = ifelse(is.na(Family), "Unassigned", Family),
    Genus = ifelse(is.na(Genus), "Unassigned", Genus)
  )

## ------------------------------------------------------------
## 2. Rank ASVs within each sample
## ------------------------------------------------------------

asv_ranked <- asv_long_tax %>%
  filter(rel_abundance > 0) %>%
  group_by(SampleID) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  ungroup()

## ------------------------------------------------------------
## 3. Identify samples where top1 is ASV_1, ASV_2, ASV_9 or ASV_10
## ------------------------------------------------------------

top1_samples <- asv_ranked %>%
  filter(rank == 1) %>%
  select(SampleID, top1_ASV = ASV_ID) %>%
  filter(top1_ASV %in% c("ASV_1", "ASV_2", "ASV_9", "ASV_10"))

## ------------------------------------------------------------
## 4. Function to summarise neighbourhoods for Family or Genus
## ------------------------------------------------------------

make_neighbourhood_table <- function(tax_level) {
  
  tax_level <- rlang::sym(tax_level)
  
  neighbourhood_summary <- asv_ranked %>%
    inner_join(top1_samples, by = "SampleID") %>%
    filter(rank >= 2, rank <= 19) %>%
    group_by(top1_ASV, SampleID, taxon = !!tax_level) %>%
    summarise(
      taxon_abundance_ranks_2_19 = sum(rel_abundance),
      .groups = "drop"
    ) %>%
    group_by(top1_ASV, taxon) %>%
    summarise(
      mean_abundance = mean(taxon_abundance_ranks_2_19),
      .groups = "drop"
    ) %>%
    group_by(top1_ASV) %>%
    arrange(desc(mean_abundance), .by_group = TRUE) %>%
    slice_head(n = 10) %>%
    ungroup()
  
  neighbourhood_table <- neighbourhood_summary %>%
    mutate(
      percent = round(100 * mean_abundance, 2)
    ) %>%
    select(taxon, top1_ASV, percent) %>%
    pivot_wider(
      names_from = top1_ASV,
      values_from = percent
    ) %>%
    arrange(taxon)
  
  list(
    summary = neighbourhood_summary,
    table = neighbourhood_table
  )
}

## ------------------------------------------------------------
## 5. Family neighbourhood table
## ------------------------------------------------------------

family_neighbourhood <- make_neighbourhood_table("Family")

family_neighbourhood_summary <- family_neighbourhood$summary
family_neighbourhood_table <- family_neighbourhood$table

print(family_neighbourhood_table, n = Inf)

write.csv(
  family_neighbourhood_summary,
  "family_neighbourhood_top1_ASV_1_2_9_10_ranks_2_19_summary.csv",
  row.names = FALSE
)

write.csv(
  family_neighbourhood_table,
  "family_neighbourhood_top1_ASV_1_2_9_10_ranks_2_19_table.csv",
  row.names = FALSE
)

## ------------------------------------------------------------
## 6. Genus neighbourhood table
## ------------------------------------------------------------

genus_neighbourhood <- make_neighbourhood_table("Genus")

genus_neighbourhood_summary <- genus_neighbourhood$summary
genus_neighbourhood_table <- genus_neighbourhood$table

print(genus_neighbourhood_table, n = Inf)

write.csv(
  genus_neighbourhood_summary,
  "genus_neighbourhood_top1_ASV_1_2_9_10_ranks_2_19_summary.csv",
  row.names = FALSE
)

write.csv(
  genus_neighbourhood_table,
  "genus_neighbourhood_top1_ASV_1_2_9_10_ranks_2_19_table.csv",
  row.names = FALSE
)

```


```{r pearson-correlation-intra-inter-combinations, message=FALSE, warning=FALSE}


library(phyloseq)
library(dplyr)
library(tibble)
library(ggplot2)
library(cowplot)

## ASV abundance matrix: samples as rows, ASVs as columns
otu_mat <- as(otu_table(ps), "matrix")

if (taxa_are_rows(ps)) {
  otu_mat <- t(otu_mat)
}

## Metadata with explicit SampleID column
metadata_corr <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID") %>%
  select(SampleID, Componentes) %>%
  mutate(
    Complexity = case_when(
      nchar(Componentes) == 1 ~ "Single",
      nchar(Componentes) == 2 ~ "Pair",
      nchar(Componentes) == 3 ~ "Triplet",
      TRUE ~ NA_character_
    ),
    Complexity = factor(Complexity, levels = c("Single", "Pair", "Triplet"))
  )

## Pearson correlation between samples
corr_mat <- cor(t(otu_mat), method = "pearson")

## Convert correlation matrix to pairwise table
sample_pairs <- as.data.frame(as.table(corr_mat)) %>%
  rename(sample1 = Var1, sample2 = Var2, corr = Freq) %>%
  mutate(
    sample1 = as.character(sample1),
    sample2 = as.character(sample2)
  ) %>%
  filter(sample1 < sample2) %>%
  left_join(metadata_corr, by = c("sample1" = "SampleID")) %>%
  rename(
    comb1 = Componentes,
    complexity1 = Complexity
  ) %>%
  left_join(metadata_corr, by = c("sample2" = "SampleID")) %>%
  rename(
    comb2 = Componentes,
    complexity2 = Complexity
  ) %>%
  mutate(
    comparison = ifelse(
      comb1 == comb2,
      "Intra-combination",
      "Inter-combination"
    ),
    intra_complexity = ifelse(
      comparison == "Intra-combination",
      as.character(complexity1),
      NA_character_
    ),
    intra_complexity = factor(
      intra_complexity,
      levels = c("Single", "Pair", "Triplet")
    )
  )

## Summary statistics
corr_stats <- sample_pairs %>%
  group_by(comparison) %>%
  summarise(
    n = n(),
    mean_corr = mean(corr, na.rm = TRUE),
    sd_corr = sd(corr, na.rm = TRUE),
    median_corr = median(corr, na.rm = TRUE),
    q25_corr = quantile(corr, 0.25, na.rm = TRUE),
    q75_corr = quantile(corr, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

print(corr_stats)

## Main density plot
p_corr_density <- ggplot(sample_pairs,
                         aes(x = corr,
                             fill = comparison,
                             color = comparison)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  scale_fill_manual(values = c(
    "Intra-combination" = "#4D4D4D",
    "Inter-combination" = "#BDBDBD"
  )) +
  scale_color_manual(values = c(
    "Intra-combination" = "#252525",
    "Inter-combination" = "#737373"
  )) +
  theme_bw() +
  labs(
    x = "Pearson correlation",
    y = "Density",
    title = "Within- vs between-combination Pearson correlations",
    fill = "Comparison",
    color = "Comparison"
  )

## Inset
p_inset <- sample_pairs %>%
  filter(comparison == "Intra-combination") %>%
  ggplot(aes(x = corr,
             fill = intra_complexity,
             color = intra_complexity)) +
  geom_density(alpha = 0.30, linewidth = 0.6) +
  scale_fill_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  scale_color_manual(values = c(
    "Single" = "#1b9e77",
    "Pair" = "#d95f02",
    "Triplet" = "#7570b3"
  )) +
  theme_bw(base_size = 8) +
  labs(
    x = NULL,
    y = NULL,
    title = "Intra only",
    fill = "Complexity",
    color = "Complexity"
  ) +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    plot.title = element_text(size = 9),
    axis.text = element_text(size = 7),
    panel.grid.minor = element_blank()
  )

## Combined plot
p_corr_density_inset <- ggdraw() +
  draw_plot(p_corr_density) +
  draw_plot(
    p_inset,
    x = 0.09,
    y = 0.55,
    width = 0.38,
    height = 0.38
  )

p_corr_density_inset

ggsave(
  "pearson_correlation_intra_inter_density_with_inset.png",
  plot = p_corr_density_inset,
  width = 8,
  height = 6,
  dpi = 300
)




## ------------------------------------------------------------
## 7. Final version: no title + legend inside below inset
## ------------------------------------------------------------

p_corr_density_final <- ggplot(sample_pairs,
                               aes(x = corr,
                                   fill = comparison,
                                   color = comparison)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  scale_fill_manual(values = c(
    "Intra-combination" = "#4D4D4D",
    "Inter-combination" = "#BDBDBD"
  )) +
  scale_color_manual(values = c(
    "Intra-combination" = "#252525",
    "Inter-combination" = "#737373"
  )) +
  theme_bw() +
  labs(
    x = "Pearson correlation",
    y = "Density",
    title = NULL,
    fill = "Comparison",
    color = "Comparison"
  ) +
  theme(
    legend.position = c(0.28, 0.4),
    legend.background = element_rect(
      fill = scales::alpha("white", 0.75),
      colour = NA
    ),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

p_corr_density_inset_final <- ggdraw() +
  draw_plot(p_corr_density_final) +
  draw_plot(
    p_inset,
    x = 0.09,
    y = 0.55,
    width = 0.38,
    height = 0.38
  )

p_corr_density_inset_final

ggsave(
  "pearson_correlation_intra_inter_density_with_inset_final.png",
  plot = p_corr_density_inset_final,
  width = 8,
  height = 6,
  dpi = 300
)

```

### Distance to the compositional Bray–Curtis centroid

This chunk defines the global rhizosphere attractor as the mean compositional profile across all samples. Bray–Curtis distances are then calculated from each sample to this compositional centroid. Pair and triplet communities are shown as density distributions, while each single inoculum is represented by its letter positioned at its mean distance to the centroid.

```{r}
library(phyloseq)
library(dplyr)
library(tibble)
library(ggplot2)
library(vegan)

## ------------------------------------------------------------
## 1. Extract ASV abundance matrix
## ------------------------------------------------------------

otu_mat <- as(otu_table(ps), "matrix")

## Samples as rows, ASVs as columns
if (taxa_are_rows(ps)) {
  otu_mat <- t(otu_mat)
}

## Convert to relative abundance, just to be explicit
otu_rel <- sweep(otu_mat, 1, rowSums(otu_mat), "/")

## ------------------------------------------------------------
## 2. Define compositional centroid
## ------------------------------------------------------------
## The centroid is the mean relative abundance profile across all samples.

centroid_profile <- colMeans(otu_rel)

## Add centroid as an extra pseudo-sample
otu_rel_with_centroid <- rbind(
  otu_rel,
  Global_centroid = centroid_profile
)

## ------------------------------------------------------------
## 3. Bray-Curtis distances to centroid
## ------------------------------------------------------------

bray_to_centroid <- vegan::vegdist(
  otu_rel_with_centroid,
  method = "bray"
)

bray_centroid_mat <- as.matrix(bray_to_centroid)

distance_to_centroid <- bray_centroid_mat[
  rownames(otu_rel),
  "Global_centroid"
]

## ------------------------------------------------------------
## 4. Metadata + distance to centroid
## ------------------------------------------------------------

metadata_centroid <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID") %>%
  mutate(
    Complexity = case_when(
      nchar(Componentes) == 1 ~ "Single",
      nchar(Componentes) == 2 ~ "Pair",
      nchar(Componentes) == 3 ~ "Triplet",
      TRUE ~ NA_character_
    ),
    Complexity = factor(Complexity, levels = c("Single", "Pair", "Triplet")),
    distance_to_centroid = distance_to_centroid[SampleID]
  )

save(metadata_centroid, file = "metadata_centroid.RData")
## ------------------------------------------------------------
## 5. Mean distance of each single inoculum
## ------------------------------------------------------------

single_mean_distances <- metadata_centroid %>%
  filter(Complexity == "Single") %>%
  group_by(Componentes) %>%
  summarise(
    mean_distance_to_centroid = mean(distance_to_centroid, na.rm = TRUE),
    .groups = "drop"
  )

print(single_mean_distances)

## ------------------------------------------------------------
## 6. Plot
## ------------------------------------------------------------

p_centroid_distance <- ggplot() +
  geom_density(
    data = metadata_centroid %>%
      filter(Complexity %in% c("Pair", "Triplet")),
    aes(
      x = distance_to_centroid,
      fill = Complexity,
      color = Complexity
    ),
    alpha = 0.25,
    linewidth = 0.8
  ) +
  geom_text(
    data = single_mean_distances,
    aes(
      x = mean_distance_to_centroid,
      y = 3,
      label = Componentes
    ),
    size = 5,
    fontface = "bold"
  ) +
scale_color_manual(values = c(
  "Pair" = "#d95f02",
  "Triplet" = "#7570b3"
)) +
scale_fill_manual(values = c(
  "Pair" = "#d95f02",
  "Triplet" = "#7570b3"
)) +
  labs(
    x = "Bray-Curtis distance to compositional centroid",
    y = "Density",
    #title = "Distance to the global rhizosphere compositional centroid",
    #subtitle = "Pairs and triplets as density curves; single inocula shown as mean positions",
    color = "Complexity",
    fill = "Complexity"
  ) +
theme_bw() +
  theme(
    legend.position = c(0.78, 0.80),
    legend.background = element_rect(
      fill = scales::alpha("white", 0.75),
      colour = NA
    )
  )

p_centroid_distance

ggsave(
  "bray_curtis_distance_to_compositional_centroid.png",
  plot = p_centroid_distance,
  width = 8,
  height = 6,
  dpi = 300
)
```


### ASV-level assembly trait space in 2D

This chunk uses the filtered ASV table, retaining only ASVs reaching at least 0.1% relative abundance in at least two samples. For each retained ASV, the source set is defined from the single-inoculum communities. We then calculate three assembly axes: conditional persistence, mean retention in eligible mixtures, and context dependence. The resulting ASV-level trait space is visualized as three 2D plots. Each ASV is shown by its numeric ASV label, with text size proportional to source breadth and colour indicating taxonomic family.
```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

## ------------------------------------------------------------
## 1. Filter ASVs: >=0.1% relative abundance in at least 2 samples
## ------------------------------------------------------------

asv_filter_summary <- asv_long %>%
  group_by(ASV) %>%
  summarise(
    n_samples_above_threshold = sum(rel_abundance >= 0.001),
    .groups = "drop"
  )

asvs_kept <- asv_filter_summary %>%
  filter(n_samples_above_threshold >= 2) %>%
  pull(ASV)

asv_long_filt <- asv_long %>%
  filter(ASV %in% asvs_kept)

## ------------------------------------------------------------
## 2. Add ASV_ID and taxonomy
## ------------------------------------------------------------

taxonomy_info <- asv_context_summary_tax %>%
  select(ASV, Family, Genus) %>%
  distinct() %>%
  mutate(
    Family = ifelse(is.na(Family), "Unassigned", Family),
    Genus = ifelse(is.na(Genus), "Unassigned", Genus)
  )

asv_long_filt <- asv_long_filt %>%
  left_join(asv_ids, by = "ASV") %>%
  left_join(taxonomy_info, by = "ASV") %>%
  mutate(
    ASV_number = as.numeric(gsub("ASV_", "", ASV_ID))
  )

## ------------------------------------------------------------
## 3. Mean abundance per ASV x context
## ------------------------------------------------------------

asv_context_mean <- asv_long_filt %>%
  group_by(ASV, ASV_ID, ASV_number, Family, Genus, Componentes) %>%
  summarise(
    mean_abundance = mean(rel_abundance),
    present = mean_abundance >= 0.001,
    .groups = "drop"
  ) %>%
  mutate(
    context_size = nchar(Componentes)
  )

## ------------------------------------------------------------
## 4. Define source set from singles
## ------------------------------------------------------------

source_set <- asv_context_mean %>%
  filter(context_size == 1, present) %>%
  group_by(ASV, ASV_ID, ASV_number, Family, Genus) %>%
  summarise(
    source_components = list(Componentes),
    source_breadth = n(),
    .groups = "drop"
  )

## Keep only ASVs detected in at least one single inoculum
asv_context_mean_sources <- asv_context_mean %>%
  inner_join(
    source_set,
    by = c("ASV", "ASV_ID", "ASV_number", "Family", "Genus")
  )

## ------------------------------------------------------------
## 5. Evaluate only eligible mixtures
## ------------------------------------------------------------

eligible_mixtures <- asv_context_mean_sources %>%
  filter(context_size %in% c(2, 3)) %>%
  rowwise() %>%
  mutate(
    mixture_components = list(strsplit(Componentes, "")[[1]]),
    eligible = length(
      intersect(
        unlist(source_components),
        unlist(mixture_components)
      )
    ) > 0
  ) %>%
  ungroup() %>%
  filter(eligible)

## ------------------------------------------------------------
## 6. Add best source-single abundance for each eligible mixture
## ------------------------------------------------------------

single_abundances <- asv_context_mean_sources %>%
  filter(context_size == 1) %>%
  select(
    ASV,
    single_component = Componentes,
    single_abundance = mean_abundance
  )

eligible_mixtures <- eligible_mixtures %>%
  rowwise() %>%
  mutate(
    best_source_single_abundance = max(
      single_abundances$single_abundance[
        single_abundances$ASV == ASV &
          single_abundances$single_component %in% unlist(mixture_components)
      ],
      na.rm = TRUE
    )
  ) %>%
  ungroup()

## ------------------------------------------------------------
## 7. Calculate retention per ASV x eligible mixture
## ------------------------------------------------------------

epsilon <- 1e-6

eligible_mixtures <- eligible_mixtures %>%
  mutate(
    retention = log2(
      (mean_abundance + epsilon) /
        (best_source_single_abundance + epsilon)
    )
  )

## ------------------------------------------------------------
## 8. ASV-level assembly axes
## ------------------------------------------------------------

asv_trait_space <- eligible_mixtures %>%
  group_by(ASV, ASV_ID, ASV_number, Family, Genus, source_breadth) %>%
  summarise(
    n_eligible_mixtures = n(),
    conditional_persistence = mean(present),
    mean_retention = mean(retention, na.rm = TRUE),
    context_dependence = sd(retention, na.rm = TRUE),
    max_abundance_in_mixtures = max(mean_abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    context_dependence = ifelse(is.na(context_dependence), 0, context_dependence)
  )

print(asv_trait_space, n = 30)

write.csv(
  asv_trait_space,
  "asv_assembly_trait_space.csv",
  row.names = FALSE
)

## ------------------------------------------------------------
## 9. Collapse taxonomy to top 10 families + Other
## ------------------------------------------------------------

top10_families <- asv_trait_space %>%
  group_by(Family) %>%
  summarise(
    total_weight = sum(max_abundance_in_mixtures, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_weight)) %>%
  slice_head(n = 10) %>%
  pull(Family)

asv_trait_space_plot <- asv_trait_space %>%
  mutate(
    Family_plot = ifelse(Family %in% top10_families, Family, "Other")
  )

## palette
family_levels <- c(sort(setdiff(unique(asv_trait_space_plot$Family_plot), "Other")), "Other")

family_colors <- setNames(
  c(scales::hue_pal()(length(family_levels)-1), "grey75"),
  family_levels
)

## ------------------------------------------------------------
## 10. Plot 1: retention vs persistence
## ------------------------------------------------------------

p1 <- ggplot(
  asv_trait_space_plot,
  aes(
    x = mean_retention,
    y = conditional_persistence,
    size = source_breadth,
    color = Family_plot
  )
) +
  geom_point(alpha = 0.75) +
  scale_color_manual(values = family_colors) +
  theme_bw() +
  labs(
    x = "Mean retention",
    y = "Conditional persistence",
    #title = "Retention vs persistence",
    color = "Family",
    size = "Source breadth"
  )

## ------------------------------------------------------------
## 11. Plot 2: retention vs context dependence
## ------------------------------------------------------------

p2 <- ggplot(
  asv_trait_space_plot,
  aes(
    x = mean_retention,
    y = context_dependence,
    size = source_breadth,
    color = Family_plot
  )
) +
  geom_point(alpha = 0.75) +
  scale_color_manual(values = family_colors) +
  theme_bw() +
  labs(
    x = "Mean retention",
    y = "Context dependence",
    #title = "Retention vs context dependence",
    color = "Family",
    size = "Source breadth"
  )

## ------------------------------------------------------------
## 12. Combine with shared legend
## ------------------------------------------------------------

p_combined <- (p1 + p2) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p_combined

ggsave(
  "asv_trait_space_2plots.png",
  plot = p_combined,
  width = 12,
  height = 6,
  dpi = 300
)

```






