---
title: "Script capítulo Coalescencia"
author: "Ruben Chaboy"
date: 
output: html_document
---
<style>
hr {
  border: 0;
  height: 8px;        
  background: #000;   
  margin: 20px 0;     
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


---

## Loading of the libraries and data 

Here we load the libraries we will use in the study. The data generated from sequence processing following the standard workflow of the DADA2 package are loaded. These data contain the table of ASV abundances per sample, as well as their taxonomic assignment. 


```{r}
knitr::opts_knit$set(root.dir = getwd())
suppressMessages(suppressWarnings({
datos <- "dada2multithreads.RData"
suppressPackageStartupMessages({
library(dada2)
library(ggplot2)
library(here)
library(phyloseq)
library(microbiome)
library(dplyr)
library(writexl)
library(readxl)
library(openxlsx)
library(gridExtra)
library(patchwork)
library(Biostrings)
library(tidyr)
library(tibble)
library(vegan)
library(scales)
library(car)})
library(emmeans)
library(officer)
library(flextable)
library(patchwork)
library(broom)
library(pheatmap)
library(picante)
library(reshape2)
library(ppcor)
library(gtsummary)
library(stringr)
library(readr)
library(eulerr)
library(rlang)
library(gtools)
library(ragg)
library(magick)
library(igraph)
library(ggraph)
library(tidygraph)
theme_set(theme_bw())
load(here("data",datos)) })) 

dir.create("test")
dir.create("graficos")
dir.create("graficos/individual")
dir.create("graficos/parejas")

out_graf <- "graficos"
out_graf_1 <- "graficos/individual"
out_graf_2 <- "graficos/parejas"

if (!dir.exists(out_graf_1)) dir.create(out_graf_1, recursive = TRUE)
if (!dir.exists(out_graf_2)) dir.create(out_graf_2, recursive = TRUE)

```

## Data preparation
  
1. All experimental data were imported from an Excel file.
2. A phyloseq object was created.

```{r}

samdf <- read_excel("metadata_Coales.xlsx")
samdf <- as.data.frame(samdf)


rownames(samdf) <- samdf$Nombre

ps <- phyloseq(otu_table(seqtab_nochim, taxa_are_rows = FALSE), 
               sample_data(samdf), 
               tax_table(tax))

```

### Eliminación de controles y muestras no representativas {#controles}

Samples that correspond to controls, DNA extraction controls (Cext), and negative PCR controls (Cneg) are eliminated. Samples containing community B are also eliminated. Finally, samples from dead or small plants are eliminated.
```{r}

ps_nocontrol <- subset_samples(
  ps,
  !(Tipo %in% c("Cext", "Cneg")) &
  !grepl("B", Componentes) &
   !Observaciones %in% c("muerta", "pequeña")
)

```

### Removal of non-bacterial sequences

Sequences corresponding to mitochondria, chloroplasts, and eukaryotes were removed. This was done because the selected primers may also amplify the 16S region of these organelles.


```{r}

print("Inicial:")
ps_nocontrol
inicial<-sum(otu_table(ps_nocontrol))
print ("Aqui vamos a quitar cloroplastos, mitocondrias y eucariotas")
ps_nocontrol_filtrado<- subset_taxa(ps_nocontrol, Class!="Chloroplast")
sin_cloroplastos<-sum(otu_table(ps_nocontrol_filtrado))
ps_nocontrol_filtrado <- subset_taxa(ps_nocontrol_filtrado, Family!="Mitochondria")
sin_mitocondrias<-sum(otu_table(ps_nocontrol_filtrado))
ps_nocontrol_filtrado <- subset_taxa(ps_nocontrol_filtrado, Kingdom!="Eukaryota")
sin_eucariotas<-sum(otu_table(ps_nocontrol_filtrado))
valores<-c(inicial,sin_cloroplastos,sin_mitocondrias,sin_eucariotas)
print("inicial,sin_cloroplastos,sin_mitocondrias,sin_eucariotas")
print(valores)
print(paste0("Total inicial= ",inicial))
print(paste0("  quitando cloroplastos= ",sin_cloroplastos))
print(paste0("  quitando cloroplastos y mitocondrias= ",sin_mitocondrias))
print(paste0("  quitando cloroplastos, mitocondrias y eucariotas= ",sin_eucariotas))
print("Filtrado:")
ps_nocontrol_filtrado

suppressMessages(summarize_phyloseq(ps_nocontrol_filtrado))

```


### Correction of NAs 

Samples with MPN values recorded as NA, either because the corresponding bacteria did not grow in those samples or because the values were below the detection limit (<110000 MPN/mL), were converted to the detection-limit value of 220000 MPN per plant.

1. The text values ‘NA’ are replaced with missing values, which are coded as NA, but not in text format as they appear in the original data sheet. 
2. The MPN column was converted to numeric type.
The NA (missing) values in the MPN column are replaced with 220000.
4. All values below 110000 in the MPN column were converted to 220000.

For dry weight, samples with NA values (because the aboveground part did not grow sufficiently) were assigned a value of 0 g.

1. The text values ‘NA’ are replaced with NA (missing) values.
2. The Weight column was converted to numeric type.
3.  The NA (missing) values in the Weight column are replaced with 0.


```{r}

sample_data(ps_nocontrol_filtrado)$MPN[sample_data(ps_nocontrol_filtrado)$MPN == "NA"] <- NA
sample_data(ps_nocontrol_filtrado)$MPN <- as.numeric(sample_data(ps_nocontrol_filtrado)$MPN)
sample_data(ps_nocontrol_filtrado)$MPN[is.na(sample_data(ps_nocontrol_filtrado)$MPN)] <- 220000
sample_data(ps_nocontrol_filtrado)$MPN[sample_data(ps_nocontrol_filtrado)$MPN < 220000] <- 110000

sample_data(ps_nocontrol_filtrado)$Peso_Seco[sample_data(ps_nocontrol_filtrado)$Peso_Seco == "NA"] <- NA
sample_data(ps_nocontrol_filtrado)$Peso_Seco <- as.numeric(sample_data(ps_nocontrol_filtrado)$Peso_Seco)
sample_data(ps_nocontrol_filtrado)$Peso_Seco[is.na(sample_data(ps_nocontrol_filtrado)$Peso_Seco)] <- 0

```

## Subsampling 

The number of sequences per sample was visualized to determine the subsampling depth. A fixed seed of 134 was used.

```{r}

sort(sample_sums(ps_nocontrol_filtrado), decreasing = TRUE)

suppressMessages(suppressWarnings({
  set.seed(134)  
  ps_nocontrol_filtrado_rare <- rarefy_even_depth(ps_nocontrol_filtrado, sample.size = 23394, trimOTUs = TRUE)
}))

png("graficos/profundidad_secuenciacion.png", width = 800, height = 600)

plot(
  sort(sample_sums(ps_nocontrol_filtrado)),
  ylab = "Sequencing depth",
  xlab = "Samples ordered by depth",
  main = "Sequencing depth per sample",
  cex.lab = 1.6,   
  cex.axis = 1.4,   
  cex.main = 1.8    
)
abline(h = 23394, col = "red", lwd = 2)

dev.off()

plot(
  sort(sample_sums(ps_nocontrol_filtrado)),
  ylab = "Sequencing depth",
  xlab = "Samples ordered by depth",
  main = "Sequencing depth per sample",
  cex.lab = 1.6,   
  cex.axis = 1.4,   
  cex.main = 1.8    
)

```

## Saving and loading 

Saving of the phyloseq object and loading of data with phylogenetic distance.


```{r}

saveRDS(ps_nocontrol_filtrado_rare, "ps_nocontrol_filtrado_rare.rds")

```

In a separate script, the sequences from the previously generated phyloseq object were matched against a reference database with an associated phylogenetic tree and taxonomy (e.g., Greengenes2). This allowed the phylogenetic tree to be incorporated into the object and enabled analyses based on phylogenetic metrics (e.g., Faith's PD, UniFrac, etc.).

```{r}

nuevosdatos <- "phyloseq_objects_with_tree.Rdata"

load(here("data",nuevosdatos ))

```

---

# **Diversity analysis and correlations** 

## Creation of working variables 

```{r}

datos_est <- as(sample_data(ps_nocontrol_filtrado_rare), "data.frame")
datos_est$log_MPN <- log10(datos_est$MPN)

Riqueza <- estimate_richness(ps_nocontrol_filtrado_rare, measures = "Observed")
Homogeneidad <- estimate_richness(ps_nocontrol_filtrado_rare, measures = "Shannon")

Faith_D <- pd(as(otu_table(new_ps), "matrix"), phy_tree(new_ps))
datos_est$Riqueza <- Riqueza[rownames(datos_est), "Observed"]
datos_est$Homogeneidad <- Homogeneidad[rownames(datos_est), "Shannon"]
datos_est$Faith_D <- Faith_D[rownames(datos_est), "PD"]

# Variable type checking

datos_est$Componentes   <- factor(datos_est$Componentes)  
datos_est$Tipo   <- factor(datos_est$Tipo) 


# Data preparation for subsequent analysis”
abundancia_rel <- transform_sample_counts(ps_nocontrol_filtrado_rare, function(x) x / sum(x))
abundancia_rel_df <- as.data.frame(otu_table(abundancia_rel))

abund_t <- t(abundancia_rel_df)
abund_t <- abund_t %>%
  as.data.frame() %>%
  tibble::rownames_to_column("clade_name")

abund_t <- as_tibble(abund_t)

# create new IDs ‘T_Coal_001’, ‘T_Coal_002...
n <- nrow(abund_t)
width <- max(3, nchar(as.character(n)))                       
new_ids <- sprintf(paste0("T_Coal_%0", width, "d"), seq_len(n))

# 2) preserve the original sequence and replace clade_name with the ID
abund_t <- abund_t %>%
  mutate(
    original_seq = as.character(clade_name),   
    clade_name = new_ids                       
  )

```

```{r}

table(sample_data(ps_nocontrol_filtrado_rare)$Componentes)

```


## **AMicrobial Composition Analysis ** 

## Microbial Composition - Bray-Curtis  


In this section, a loop is first created to generate all the phyloseq objects for every pairwise combination. Each object contains the samples from the individual components as well as those from the corresponding pair itself.

Using a second loop, the microbial composition of each of the previous objects is represented through NMDS analysis based on Bray–Curtis distances.

```{r}

Suelos <- c("A", "C", "D", "E", "F", "G", "H")
pares <- combn(Suelos, 2, simplify = FALSE)

# Create and save individual plots
all_plots <- list()

for (par in pares) {
  comp1 <- par[1]
  comp2 <- par[2]
  Pareja <- paste0(comp1, comp2)
  obj_pareja <- paste0("ps_nocontrol_filtrado_rare", Pareja)
  
  ps_filtrado_pareja <- subset_samples(
    ps_nocontrol_filtrado_rare,
    Componentes %in% c(comp1, comp2, Pareja)
  )
  assign(obj_pareja, ps_filtrado_pareja)
}

for (par in pares) {
  comp1 <- par[1]
  comp2 <- par[2]
  
  pareja <- paste0(comp1, comp2)
  ps_filtrado_pareja <- get(paste0("ps_nocontrol_filtrado_rare", pareja))
  ps_prop_pareja <- transform_sample_counts(ps_filtrado_pareja, function(otu) otu / sum(otu))
  
set.seed(123)
  ord_nmds_bray_pareja <- ordinate(ps_prop_pareja, method = "NMDS", distance = "bray")
  assign(paste0("ps_prop_", pareja), ps_prop_pareja)
  assign(paste0("ord_nmds_bray_", pareja), ord_nmds_bray_pareja)
}

for (par in pares) {
  comp1 <- par[1]
  comp2 <- par[2]
  pareja <- paste0(comp1, comp2)
  
  ps_prop_pareja <- get(paste0("ps_prop_", pareja))
  ord_nmds_bray_pareja <- get(paste0("ord_nmds_bray_", pareja))
  

p_pareja <- plot_ordination(
  ps_prop_pareja, ord_nmds_bray_pareja, 
  shape = "Tipo", color = "Componentes", 
  title = paste("NMDS Bray-Curtis")
)
    p_pareja$layers <- p_pareja$layers[-1]

 
p_pareja <- p_pareja +
  geom_point(
    aes(fill = Componentes, shape = Tipo),
    size = 4,
    stroke = 0.8,
    color = "black",
    position = position_jitter(width = 0.09, height = 0.09)
  ) +
  scale_shape_manual(values = c(Individual = 22, Pareja = 21)) +
  scale_fill_brewer(palette = "Set1") +   # o Set2 / Dark2
  guides(
    fill  = guide_legend(override.aes = list(shape = 21, color = "black")),
    shape = guide_legend(override.aes = list(fill = "white", color = "black"))
  ) +
  theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      axis.text.x = element_text(angle = 0, hjust = 1, size = 10),
      axis.text.y = element_text(size = 10),
      legend.title = element_text(margin = margin(b = 10)))
  
  # Guardar plot individual
  ggsave(filename = paste0("graficos/parejas/Bray-Curtis/NMDS_", pareja, ".png"), 
         plot = p_pareja, width = 7, height = 5)
  
  # Guardar en lista para paneles
  all_plots[[pareja]] <- p_pareja
}

```

## Microbial Composition - Boxplots - Bray–Curtis

For each combination of communities, Bray–Curtis distances are calculated to evaluate:

the variability within each community (intra-group).
the difference between each community and its mixture (inter-group).

Finally, the results are represented using boxplots.

```{r}

Suelos <- c("A", "C", "D", "E", "F", "G", "H")
pares <- combn(Suelos, 2, simplify = FALSE)

out_dir_plots <- "graficos/parejas/cajas_BrayCurtis"
out_dir_tables <- "parejas/bray"
dir.create(out_dir_plots, recursive = TRUE, showWarnings = FALSE)
dir.create(out_dir_tables, recursive = TRUE, showWarnings = FALSE)

# Change the labels if necessary
GROUP_COL <- "Componentes"
MIX_COL <- "Tipo"
INDIV_LABEL <- "Individual"
PAIR_LABEL <- "Pareja"
TRANSFORM_TO_PROPORTIONS <- TRUE

# -----------------------------
# Function setup
# -----------------------------
.dist_values <- function(dist_mat, ids1, ids2 = ids1) {
  ids1 <- intersect(ids1, rownames(dist_mat))
  ids2 <- intersect(ids2, colnames(dist_mat))
  if (length(ids1) == 0 || length(ids2) == 0) return(numeric(0))

  if (identical(ids1, ids2)) {
    sub <- dist_mat[ids1, ids1, drop = FALSE]
    return(sub[upper.tri(sub, diag = FALSE)])
  } else {
    return(as.numeric(dist_mat[ids1, ids2, drop = FALSE]))
  }
}

# -----------------------------
# Calculate Bray-Curtis distances by blocks for an XY pair (default function)
# -----------------------------
calcular_distancias_entre_grupos <- function(ps_objeto, comp1, comp2, pareja,
                                             group_col = "Componentes",
                                             mix_col = "Tipo",
                                             indiv_label = "Individual",
                                             pair_label = "Pareja",
                                             transform_to_proportions = TRUE) {
  ps_use <- ps_objeto

  if (transform_to_proportions) {
    ps_use <- transform_sample_counts(ps_use, function(x) if (sum(x) > 0) x / sum(x) else x)
  }

  # Bray-Curtis distances
  dist_df <- as.matrix(phyloseq::distance(ps_use, method = "bray"))

  md <- data.frame(sample_data(ps_use), stringsAsFactors = FALSE) %>%
    tibble::rownames_to_column("SampleID")

  if (!group_col %in% colnames(md)) {
    stop("No existe columna '", group_col, "' en sample_data(). Columnas: ",
         paste(colnames(md), collapse = ", "))
  }
  if (!mix_col %in% colnames(md)) {
    stop("No existe columna '", mix_col, "' en sample_data(). Columnas: ",
         paste(colnames(md), collapse = ", "))
  }

  md[[group_col]] <- as.character(md[[group_col]])
  md[[mix_col]] <- as.character(md[[mix_col]])

  # Sample IDs by block
  m1 <- md %>%
    filter(.data[[group_col]] == comp1, .data[[mix_col]] == indiv_label) %>%
    pull(SampleID)

  m2 <- md %>%
    filter(.data[[group_col]] == comp2, .data[[mix_col]] == indiv_label) %>%
    pull(SampleID)

  mp <- md %>%
    filter(.data[[group_col]] == pareja, .data[[mix_col]] == pair_label) %>%
    pull(SampleID)

  out <- list()

  # intragroup comp1
  v <- .dist_values(dist_df, m1)
  if (length(v) > 0) out[[length(out) + 1]] <- tibble(
    Pareja = pareja,
    Grupo = paste0(comp1, " intra"),
    Distancia = v
  )

  # comp1 vs pair
  v <- .dist_values(dist_df, m1, mp)
  if (length(v) > 0) out[[length(out) + 1]] <- tibble(
    Pareja = pareja,
    Grupo = paste0(comp1, " vs ", pareja),
    Distancia = v
  )

  # intragroup comp2
  v <- .dist_values(dist_df, m2)
  if (length(v) > 0) out[[length(out) + 1]] <- tibble(
    Pareja = pareja,
    Grupo = paste0(comp2, " intra"),
    Distancia = v
  )

  # comp2 vs pair
  v <- .dist_values(dist_df, m2, mp)
  if (length(v) > 0) out[[length(out) + 1]] <- tibble(
    Pareja = pareja,
    Grupo = paste0(comp2, " vs ", pareja),
    Distancia = v
  )

  if (length(out) == 0) return(NULL)
  bind_rows(out)
}

# ============================================================
# Final step: for each pair, calculate distances and save the graph 
# ============================================================
all_dist <- list()

for (par in pares) {
  comp1 <- par[1]
  comp2 <- par[2]
  pareja <- paste0(comp1, comp2)

  message("Procesando pareja: ", pareja)

  obj_name <- paste0("ps_nocontrol_filtrado_rare", pareja)
  if (!exists(obj_name, inherits = TRUE)) {
    message("  ✗ No existe objeto: ", obj_name, " (saltando)")
    next
  }

  ps_filtrado_pareja <- get(obj_name, inherits = TRUE)

  ps_filtrado_pareja <- prune_taxa(taxa_sums(ps_filtrado_pareja) > 0, ps_filtrado_pareja)
  ps_filtrado_pareja <- prune_samples(sample_sums(ps_filtrado_pareja) > 0, ps_filtrado_pareja)
  
# The data to use are selected; otherwise, the default ones will be used 
  
  dist_data <- calcular_distancias_entre_grupos(
    ps_objeto = ps_filtrado_pareja,
    comp1 = comp1,
    comp2 = comp2,
    pareja = pareja,
    group_col = GROUP_COL,
    mix_col = MIX_COL,
    indiv_label = INDIV_LABEL,
    pair_label = PAIR_LABEL,
    transform_to_proportions = TRANSFORM_TO_PROPORTIONS
  )

  if (is.null(dist_data) || nrow(dist_data) == 0) {
    message("  ✗ No hay suficientes datos para ", pareja)
    next
  }

  all_dist[[pareja]] <- dist_data

  #  boxes are always ordered the same way
  orden_grupos <- c(
    paste0(comp1, " intra"),
    paste0(comp1, " vs ", pareja),
    paste0(comp2, " intra"),
    paste0(comp2, " vs ", pareja)
  )
  orden_grupos <- orden_grupos[orden_grupos %in% unique(dist_data$Grupo)]
dist_data$Grupo <- factor(dist_data$Grupo, levels = orden_grupos)

colores_cajas <- c()
for (grupo in orden_grupos) {
  if (grepl("intra", grupo)) {
    if (grepl(paste0("^", comp1), grupo)) {
      colores_cajas <- c(colores_cajas, "#E74C3C")  # Rojo (comp1 intra)
    } else {
      colores_cajas <- c(colores_cajas, "#3498DB")  # Azul (comp2 intra)
    }
  } else {
    if (grepl(paste0("^", comp1, " vs"), grupo)) {
      colores_cajas <- c(colores_cajas, "#F39C12")  # Naranja (comp1 vs pareja)
    } else {
      colores_cajas <- c(colores_cajas, "#9B59B6")  # Morado (comp2 vs pareja)
    }
  }
}
names(colores_cajas) <- orden_grupos

  p_box <- ggplot(dist_data, aes(x = Grupo, y = Distancia, fill = Grupo)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.5, size = 2, aes(color = Grupo)) +
    scale_fill_manual(values = colores_cajas, name = "Comparación") +  #añadido nuevo
     scale_color_manual(values = colores_cajas, guide = "none") + #añadido
    labs(
      title = paste("Análisis intra e inter Grupo (Distancias Bray-Curtis)"),
      x = "Comparación de componentes",
      y = "Distancia Bray–Curtis"
    ) +
    theme_minimal() +
 theme(
  plot.margin = margin(2, 2, 2, 2, unit = "pt"),
  axis.title.y = element_text(size = 14),
  axis.title.x = element_text(size = 14),
  plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
  axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
  axis.text.y = element_text(size = 14),
  legend.position = "none"
)

  ggsave(
    filename = file.path(out_dir_plots, paste0("Boxplot_", pareja, ".png")),
    plot = p_box, width = 10, height = 6, dpi = 300
  )

  message("  ✓ Guardado: ", file.path(out_dir_plots, paste0("Boxplot_", pareja, ".png")))
}

```

## Microbial Composition - Unifrac

In this section, a loop is first created to generate all phyloseq objects for every pairwise combination. Each object will contain the samples from the individual members and from the pair itself.  
The second loop is then used to represent, for each of the objects above, the microbial composition using NMDS analysis with UniFrac distances.

```{r}

Suelos <- c("A", "C", "D", "E", "F", "G", "H")
pares <- combn(Suelos, 2, simplify = FALSE)

# Create subsets of new_ps for UniFrac
for (par in pares) {
  comp1 <- par[1]
  comp2 <- par[2]
  Pareja <- paste0(comp1, comp2)
  
  ps_unifrac_pareja <- subset_samples(
    new_ps,
    Componentes %in% c(comp1, comp2, Pareja)
  )
  assign(paste0("ps_unifrac_", Pareja), ps_unifrac_pareja)
}

# Create and save individual plots with Weighted UniFrac
all_plots_unifrac_w <- list()

for (par in pares) {
  comp1 <- par[1]
  comp2 <- par[2]
  pareja <- paste0(comp1, comp2)
  
  ps_unifrac_pareja <- get(paste0("ps_unifrac_", pareja))
  set.seed(123)
  
  ord_unifrac_w_pareja <- ordinate(ps_unifrac_pareja, method = "NMDS", distance = "unifrac", weighted = TRUE)
  

  p_pareja <- plot_ordination(
  ps_unifrac_pareja, ord_unifrac_w_pareja, 
  shape = "Tipo", color = "Componentes", 
  title = paste("NMDS Unifrac")
)
    p_pareja$layers <- p_pareja$layers[-1]


 p_pareja <- p_pareja +
  geom_point(
    aes(fill = Componentes, shape = Tipo),
    size = 4,
    stroke = 0.8,
    color = "black",
    position = position_jitter(width = 0.09, height = 0.09)
  ) +
  scale_shape_manual(values = c(Individual = 22, Pareja = 21)) +
  scale_fill_brewer(palette = "Set1") +   # o Set2 / Dark2
  guides(
    fill  = guide_legend(override.aes = list(shape = 21, color = "black")),
    shape = guide_legend(override.aes = list(fill = "white", color = "black"))
  ) +
  theme_minimal() +
        theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      axis.text.x = element_text(angle = 0, hjust = 1, size = 10),
      axis.text.y = element_text(size = 10),
      legend.title = element_text(margin = margin(b = 10)))
  
  # Save individual plot
  ggsave(filename = paste0("graficos/parejas/Unifrac/Weighted_Unifrac_NMDS_", pareja, ".png"), 
         plot = p_pareja, width = 7, height = 5)
  
   # Save in a list for panels
  all_plots_unifrac_w[[pareja]] <- p_pareja
  
   # Save the ordination object
  assign(paste0("ord_unifrac_w_", pareja), ord_unifrac_w_pareja)
  
  message("Guardado: Weighted_Unifrac_NMDS_", pareja, ".png")
}

```

## Microbial Composition - Boxplots - UniFrac 

For each community combination, UniFrac distances are calculated to assess:

- the variability within each community (intra-group)
- the difference between each community and its mixture (inter-group)

Finally, the results are displayed using boxplots.

```{r}

Suelos <- c("A", "C", "D", "E", "F", "G", "H")
pares <- combn(Suelos, 2, simplify = FALSE)

out_dir <- "graficos/parejas/cajas_unifrac"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# Change the labels if necessary

GROUP_COL <- "Componentes"
MIX_COL <- "Tipo"
INDIV_LABEL <- "Individual"  
PAIR_LABEL  <- "Pareja"      

# -----------------------------
# Helper function: calculate distances within and between groups of samples
# -----------------------------
.dist_values <- function(dist_mat, ids1, ids2 = ids1) {
  ids1 <- intersect(ids1, rownames(dist_mat))
  ids2 <- intersect(ids2, colnames(dist_mat))
  if (length(ids1) == 0 || length(ids2) == 0) return(numeric(0))

  if (identical(ids1, ids2)) {
    sub <- dist_mat[ids1, ids1, drop = FALSE]
    return(sub[upper.tri(sub, diag = FALSE)])
  } else {
    return(as.numeric(dist_mat[ids1, ids2, drop = FALSE]))
  }
}

# -----------------------------
# Calculate UniFrac distances by blocks for an XY pair (default function)
# -----------------------------
calcular_distancias_unifrac_weighted <- function(ps_objeto, comp1, comp2, pareja,
                                                 group_col = "Componentes",
                                                 mix_col = "Tipo",
                                                 indiv_label = "individual",
                                                 pair_label = "pareja") {
  ps_use <- ps_objeto

# remove empty samples if there are any
  ps_use <- prune_samples(sample_sums(ps_use) > 0, ps_use)

  # Weighted UniFrac
  dist_df <- as.matrix(phyloseq::distance(ps_use, method = "unifrac", weighted = TRUE))

  md <- data.frame(sample_data(ps_use), stringsAsFactors = FALSE) %>%
    tibble::rownames_to_column("SampleID")

  if (!group_col %in% colnames(md)) stop("No existe columna '", group_col, "' en sample_data().")
  if (!mix_col %in% colnames(md)) stop("No existe columna '", mix_col, "' en sample_data().")

  md[[group_col]] <- as.character(md[[group_col]])
  md[[mix_col]] <- as.character(md[[mix_col]])

  m1 <- md %>% filter(.data[[group_col]] == comp1,  .data[[mix_col]] == indiv_label) %>% pull(SampleID)
  m2 <- md %>% filter(.data[[group_col]] == comp2,  .data[[mix_col]] == indiv_label) %>% pull(SampleID)
  mp <- md %>% filter(.data[[group_col]] == pareja, .data[[mix_col]] == pair_label)  %>% pull(SampleID)

  out <- list()

    # intragroup comp1
  v <- .dist_values(dist_df, m1)
  if (length(v) > 0) out[[length(out) + 1]] <- tibble(Grupo = paste0(comp1, " intra"), Distancia = v)

    # comp1 vs pair

  v <- .dist_values(dist_df, m1, mp)
  if (length(v) > 0) out[[length(out) + 1]] <- tibble(Grupo = paste0(comp1, " vs ", pareja), Distancia = v)

    # intragroup comp2

  v <- .dist_values(dist_df, m2)
  if (length(v) > 0) out[[length(out) + 1]] <- tibble(Grupo = paste0(comp2, " intra"), Distancia = v)

    # comp2 vs pair

  v <- .dist_values(dist_df, m2, mp)
  if (length(v) > 0) out[[length(out) + 1]] <- tibble(Grupo = paste0(comp2, " vs ", pareja), Distancia = v)

  if (length(out) == 0) return(NULL)
  bind_rows(out)
}

# ============================================================
# Final step: for each pair, calculate distances and save the graph 
# ============================================================
for (par in pares) {
  comp1 <- par[1]
  comp2 <- par[2]
  pareja <- paste0(comp1, comp2)

  message("Procesando pareja: ", pareja)

 # Subset from the tree-containing object (new_ps)
  ps_pair <- subset_samples(
    new_ps,
    Componentes %in% c(comp1, comp2, pareja)
  )

# (optional) if any taxa remain without counts after filtering
  ps_pair <- prune_taxa(taxa_sums(ps_pair) > 0, ps_pair)
  ps_pair <- prune_samples(sample_sums(ps_pair) > 0, ps_pair)

  dist_data <- calcular_distancias_unifrac_weighted(
    ps_objeto = ps_pair,
    comp1 = comp1,
    comp2 = comp2,
    pareja = pareja,
    group_col = GROUP_COL,
    mix_col = MIX_COL,
    indiv_label = INDIV_LABEL,
    pair_label = PAIR_LABEL
  )

  if (is.null(dist_data) || nrow(dist_data) == 0) {
    message("  ✗ No hay suficientes datos para ", pareja)
    next
  }

  # boxes are always ordered the same way
  orden_grupos <- c(
    paste0(comp1, " intra"),
    paste0(comp1, " vs ", pareja),
    paste0(comp2, " intra"),
    paste0(comp2, " vs ", pareja)
  )
  orden_grupos <- orden_grupos[orden_grupos %in% unique(dist_data$Grupo)]
  dist_data$Grupo <- factor(dist_data$Grupo, levels = orden_grupos)

  # Colours (red/blue/orange/purple)
  colores_cajas <- c()
  for (grupo in orden_grupos) {
    if (grepl("intra", grupo)) {
      if (grepl(paste0("^", comp1), grupo)) {
        colores_cajas <- c(colores_cajas, "#E74C3C")  
      } else {
        colores_cajas <- c(colores_cajas, "#3498DB")  
      }
    } else {
      if (grepl(paste0("^", comp1, " vs"), grupo)) {
        colores_cajas <- c(colores_cajas, "#F39C12")  
      } else {
        colores_cajas <- c(colores_cajas, "#9B59B6")  
      }
    }
  }
  names(colores_cajas) <- orden_grupos

  p_box <- ggplot(dist_data, aes(x = Grupo, y = Distancia, fill = Grupo)) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.5, size = 2, aes(color = Grupo)) +
    scale_fill_manual(values = colores_cajas, name = "Comparación") +  #añadido nuevo
     scale_color_manual(values = colores_cajas, guide = "none") + #añadido
    labs(
      title = paste("Análisis intra e inter grupo (Distancias Unifrac)"),
      x = "Comparación de componentes",
      y = "Distancia UniFrac"
    ) +
    theme_minimal() +
  theme(
  plot.margin = margin(2, 2, 2, 2, unit = "pt"),
  axis.title.y = element_text(size = 14),
  axis.title.x = element_text(size = 14),
  plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
  axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
  axis.text.y = element_text(size = 14),
  legend.position = "none"
)

  ggsave(
    filename = file.path(out_dir, paste0("Boxplot_Unifrac_", pareja, ".png")),
    plot = p_box, width = 10, height = 6, dpi = 300
  )

  message("  ✓ Guardado: ", file.path(out_dir, paste0("Boxplot_Unifrac_", pareja, ".png")))
}

message("✓ Proceso completado. Boxplots guardados en: ", out_dir)

```
---

## **ASV Analysis** 

## Rainbow Plots 

Rainbow plots are generated to compare ASV abundances between the pair of communities and the individual communities that compose it. First, the function is prepared; then, the different phyloseq objects are prepared with the community combinations; and finally, the function is run in a loop for all these objects.

```{r}

# ========================================================================
# Function setup
# ========================================================================

# Modified from (https://github.com/silvtal/rainbow_function)

# Function: rainbow_comparison_plot
# Purpose: compare abundances across 3 treatments (left, center, right),
# showing absence/presence combinations.
# Main parameters:
#  - abundance_data: data.frame (rows = taxa, columns = samples)
#  - metadata: data.frame with treatment and sample name columns
#  - treatment_col: name of the treatment column in metadata
#  - center_treatment: treatment to place in the center
#  - presence_threshold: threshold to consider a taxon 'present' in a treatment

rainbow_comparison_plot <- function(
    abundance_data,
    metadata,
    treatment_col,
    sample_col = "Nombre",
    center_treatment = NULL,
    Resaltar_NoResidentes = FALSE,
    output_dir = NULL,
    show_labels = FALSE,
    label_top_abundant = 15,
    label_NoResidentes = 10,
    legend_ncol = 1,
    presence_threshold = 0,
    line_size = 0.35
) {

    cat("[rainbow_comparison_plot] inicio\n"); flush.console()

    if (!treatment_col %in% colnames(metadata)) stop("La columna de tratamiento '", treatment_col, "' no existe en metadatos")
    if (!sample_col %in% colnames(metadata)) stop("La columna de muestra '", sample_col, "' no existe en metadatos")

    treatments <- unique(metadata[[treatment_col]])
    treatments <- treatments[!is.na(treatments)]
    if (length(treatments) < 2) stop("Se necesitan ≥2 tratamientos")

    if (is.null(center_treatment)) stop("center_treatment no puede ser NULL (necesitas el tratamiento del centro).")
    if (!center_treatment %in% treatments) stop("center_treatment='", center_treatment, "' no está en treatments: ", paste(treatments, collapse = ", "))

    if (length(treatments) != 3) stop("Esta función asume exactamente 3 tratamientos (izq, centro, der). Encontrados: ", length(treatments))

  # --- Reordered to ensure center position ----
# [left, center, right]
    
    if (!is.null(center_treatment) && center_treatment %in% treatments) {
      other <- setdiff(treatments, center_treatment)
      treatments <- c(other[1], center_treatment, other[-1])
    }

    cat("Tratamientos a comparar:", paste(treatments, collapse = ", "), "\n"); flush.console()
    cat("Orden de tratamientos:", paste(treatments, collapse = " -> "), "\n"); flush.console()

  # Prepare abundance data
  # Ensure row names (taxa) are in a column called 'clade_name'
    
    if (!"clade_name" %in% colnames(abundance_data)) {
      abundance_data <- as.data.frame(abundance_data) %>% tibble::rownames_to_column("clade_name")
    }

  # --- Get sample columns by treatment ----
  # For each treatment, extract the list of sample names   
    treatment_samples <- list()
    for (trt in treatments) {
      samp <- metadata %>%
        dplyr::filter(!!rlang::sym(treatment_col) == trt) %>%
        dplyr::pull(!!rlang::sym(sample_col))
          treatment_samples[[trt]] <- unique(as.character(samp))
      cat("   ", trt, " : ", length(treatment_samples[[trt]]), " muestras\n"); flush.console()
    }
   # --- Convert NA to 0 and ensure numeric values for present sample columns ----
  # This is applied only to the abundance_data columns that appear in treatment_samples
    
    all_sample_cols <- unlist(treatment_samples)
      abundance_means <- abundance_data %>%
      dplyr::mutate(
        dplyr::across(
          dplyr::all_of(all_sample_cols[all_sample_cols %in% colnames(abundance_data)]),
          ~ as.numeric(tidyr::replace_na(.x, 0))
        )
      )
   # --- Calculate means by treatment (new columns) ----
  # Create one column per treatment with the mean abundance across its samples
    
    for (trt in treatments) {
      cols <- treatment_samples[[trt]]
      cols <- cols[cols %in% colnames(abundance_means)]
      if (length(cols) > 0) {
        abundance_means[[trt]] <- rowMeans(dplyr::select(abundance_means, dplyr::all_of(cols)), na.rm = TRUE)
      } else {
        abundance_means[[trt]] <- 0
      }
    }
    cat("[rainbow_comparison_plot] medias por tratamiento calculadas\n"); flush.console()

# calcular total_mean y filtrar taxones sin abundancia en todos los tratamientos
abundance_means <- abundance_means %>%
  dplyr::mutate(
    total_mean = rowSums(dplyr::select(., dplyr::all_of(treatments)))
  ) %>%
  dplyr::filter(total_mean > 0)

     # --- Presence by treatment (using presence_threshold) ----
    for (trt in treatments) {
      abundance_means[[paste0(trt, "_present")]] <- as.integer(abundance_means[[trt]] > presence_threshold)
    }

    t1 <- treatments[1]; t2 <- treatments[2]; t3 <- treatments[3]
    col_T1 <- rlang::sym(paste0(t1, "_present"))
    col_T2 <- rlang::sym(paste0(t2, "_present"))
    col_T3 <- rlang::sym(paste0(t3, "_present"))

    # --- Final classification based on presence combinations ----
# - In_all
# - left + center
# - right + center
# - left + right (without center)
# - Individual
    
    abundance_means <- abundance_means %>%
      dplyr::mutate(
        presence_sum = !!col_T1 + !!col_T2 + !!col_T3,
        color_group_final = dplyr::case_when(
          presence_sum == 3 ~ "In_all",
          (!!col_T2 & !!col_T3) & ! (!!col_T1) ~ paste0(t3,"_",t2), # der + centro
          (!!col_T2 & !!col_T1) & ! (!!col_T3) ~ paste0(t1,"_",t2), # izq + centro
          (!!col_T1 & !!col_T3) & ! (!!col_T2) ~ paste0(t1,"_",t3), # izq + der (sin centro)
          TRUE ~ "Individual"
        )
      )
   # --- Prepare data in long format for ggplot ----
    plot_data_long <- abundance_means %>%
      tidyr::pivot_longer(cols = dplyr::all_of(treatments), names_to = "condition", values_to = "abundance") %>%
      dplyr::mutate(
        condition = factor(condition, levels = treatments),
        condition_numeric = as.numeric(condition),
        jitter_offset = (as.numeric(factor(clade_name)) %% 7 - 3) * 0.02,
        condition_jitter = condition_numeric + jitter_offset
      )
 # --- Detect taxa present at the extremes but absent in the center:
#     they need pseudo-abundance to draw the line in the center
    taxa_need_center_pseudo <- abundance_means %>%
      dplyr::filter(.data[[paste0(t1, "_present")]] == 1 &
                    .data[[paste0(t3, "_present")]] == 1 &
                    .data[[paste0(t2, "_present")]] == 0) %>%
      dplyr::pull(clade_name)
    cat("[rainbow_comparison_plot] taxa que necesitan pseudo en center:", length(taxa_need_center_pseudo), "\n"); flush.console()

      # --- Filtering: keep points with abundance > 0 or taxa that need pseudo in the center ----
    
    plot_data_long_filtered <- plot_data_long %>%
      dplyr::filter(abundance > 0 | clade_name %in% taxa_need_center_pseudo)

      # ---Compute pseudo (minimum positive / 10) to draw a point in the center where it’s missing ----
    min_pos <- min(plot_data_long_filtered$abundance[plot_data_long_filtered$abundance > 0], na.rm = TRUE)
    if (!is.finite(min_pos) || min_pos <= 0) min_pos <- 1e-6
    pseudo <- min_pos / 10

    plot_data_long_filtered <- plot_data_long_filtered %>%
      dplyr::mutate(
        abundance_plot = dplyr::case_when(
          abundance > 0 ~ abundance,
          abundance == 0 & as.character(condition) == as.character(t2) & clade_name %in% taxa_need_center_pseudo ~ pseudo,
          TRUE ~ NA_real_
        )
      )
  # --- Point sizes (abundance range to sizes) ----

    plot_data_long_filtered <- plot_data_long_filtered %>%
      dplyr::mutate(
        point_size_plot = dplyr::case_when(
          abundance > 0 & abundance < 1e-6   ~ 1.2,
          abundance > 0 & abundance < 0.001 ~ 1.5,
          abundance > 0 & abundance < 0.01  ~ 2.5,
          abundance > 0 & abundance < 0.1   ~ 3.5,
          abundance > 0                      ~ 4.5,
          abundance == 0 & as.character(condition) == as.character(t2) & clade_name %in% taxa_need_center_pseudo ~ 2.5,
          TRUE ~ 1.0
        )
      )

    cat("Usando esquema: Rainbow (sin coloreado taxonómico)\n"); flush.console()
 # --- Color scheme: define legend levels and palette ----
# Fixed legend order (with center first, then extremes, then all/individual)
    legend_levels <- c(
      paste0(t1, "_", t2),  # left + center
      paste0(t3, "_", t2),  # right  + center
      paste0(t1, "_", t3),  # left + right(without center)
      "In_all",
      "Individual"
    )

    plot_data_long_filtered <- plot_data_long_filtered %>%
      dplyr::mutate(
        color_group_final = factor(as.character(color_group_final), levels = legend_levels)
      )
  # Assign explicit colors by group (if any are missing, add gray)
    color_scheme <- c(
      "Individual" = "gray50",
      setNames("#d95f02", paste0(t1, "_", t2)),
      setNames("#1b9e77", paste0(t3, "_", t2)),
      setNames("yellow", paste0(t1, "_", t3)),
      "In_all" = "purple"
    )

 # if, for some reason, another key appears, assign gray
    present_keys <- unique(na.omit(as.character(plot_data_long_filtered$color_group_final)))
    missing_keys <- setdiff(present_keys, names(color_scheme))
    if (length(missing_keys) > 0) {
      color_scheme <- c(color_scheme, setNames(rep("gray60", length(missing_keys)), missing_keys))
    }

      # === DEFINE SCALE AND LOGARITHMIC WITH SECONDARY AXIS ===
    p <- ggplot2::ggplot(plot_data_long_filtered, ggplot2::aes(x = condition_jitter, y = abundance_plot)) +
      ggplot2::geom_line(
        data = plot_data_long_filtered %>% dplyr::filter(!is.na(abundance_plot)),
        ggplot2::aes(group = clade_name, color = color_group_final),
        size = line_size, alpha = 0.5
      ) +
      ggplot2::geom_point(
        data = plot_data_long_filtered %>% dplyr::filter(!is.na(abundance_plot)),
        ggplot2::aes(color = color_group_final, size = point_size_plot),
        alpha = 0.85
      ) +
      ggplot2::scale_size_identity(guide = "none") +
      ggplot2::scale_color_manual(
        values = color_scheme,
        breaks = legend_levels,
        guide = ggplot2::guide_legend(title = "Presence", ncol = legend_ncol),
        na.value = "gray60"
      ) +
      ggplot2::scale_x_continuous(breaks = 1:length(treatments), labels = treatments) +
      ggplot2::scale_y_log10(labels = scales::comma_format()) +
      ggplot2::labs(
        title = "ASVs relative abundance",
        x = "Components",
        y = "Relative abundance (log10)"
      ) +
      ggplot2::theme_minimal() +
      ggplot2::theme(
        plot.title   = ggplot2::element_text(size = 22, face = "bold", hjust = 0.5), 
        axis.title   = ggplot2::element_text(size = 18),
        axis.text.x  = ggplot2::element_text(size = 16, angle = 45, hjust = 1),
        axis.text.y  = ggplot2::element_text(size = 16),
        legend.title = ggplot2::element_text(size = 18),
        legend.text  = ggplot2::element_text(size = 16),
        legend.position = "right"
      )

    if (!is.null(output_dir)) {
      dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
      fname <- paste0("rainbow_", paste(treatments, collapse = "_"), ".png")
      ggplot2::ggsave(file.path(output_dir, fname), plot = p, width = 10, height = 8, dpi = 300)
      cat("Plot guardado: ", fname, "\n"); flush.console()
    }

    cat("[rainbow_comparison_plot] fin\n"); flush.console()
    return(p)
}

# ==== PAIRS ====
letras_pool <- c("A", "C", "D", "E", "F", "G", "H")
combinaciones <- gtools::combinations(n = length(letras_pool), r = 2, v = letras_pool)
combinaciones <- apply(combinaciones, 1, paste0, collapse = "")

# ==== Execute and save ====
for (dos_letras in combinaciones) {

  datos_filtrados <- datos_est %>%
    mutate(Componentes = as.character(Componentes)) %>%
    filter(grepl(paste0("^[", dos_letras, "]+$"), Componentes)) %>%
    filter(Tipo != "trío")

# get unique singles and preserve order according to the pair
  letters_order <- strsplit(dos_letras, "")[[1]]   # ej: c("A","C")
  singles <- unique(datos_filtrados$Componentes)
  singles <- singles[nchar(singles) == 1]
  singles_ordered <- intersect(letters_order, singles)

  if (length(singles_ordered) != 2) {
    message("Saltando ", dos_letras, ": faltan singles (presentes = ", paste(singles, collapse = ","), ")")
    next
  }

  niveles_deseados <- c(singles_ordered[1], dos_letras, singles_ordered[2])

  metadata_mod <- datos_filtrados %>%
    mutate(Componentes = factor(as.character(Componentes), levels = niveles_deseados))

  rainbow_comparison_plot(
    abundance_data = abund_t,
    metadata = metadata_mod,
    treatment_col = "Componentes",
    center_treatment = dos_letras,
    Resaltar_NoResidentes = FALSE,
    show_labels = FALSE,
    output_dir = "graficos/parejas/rainbow",
    legend_ncol = 1
  )
}

```

## Euler plots {#Euler}

Euler diagrams are generated to compare the presence of ASVs between the pair of communities and the individual communities that make it up. First, the functions to be used are prepared. Then the phyloseq objects containing the samples from a pair and the individual communities that make it up are created, and finally the functions are run for these objects. 

```{r}

# ---- (1) FUNCTION: extract taxa lists by treatment
extract_taxa_lists <- function(abundance_data, metadata, treatment_col,
                               sample_col = "Nombre",
                               center_treatment = NULL,
                               presence_threshold = 0) {

  treatments <- unique(metadata[[treatment_col]])
  treatments <- treatments[!is.na(treatments)]

  # Reorder: left, center (pair), right
  if (!is.null(center_treatment) && center_treatment %in% treatments) {
    other <- setdiff(treatments, center_treatment)
    treatments <- c(other[1], center_treatment, other[-1])
  }

  # Ensure "clade_name"
  if (!"clade_name" %in% colnames(abundance_data)) {
    abundance_data <- as.data.frame(abundance_data) %>%
      tibble::rownames_to_column("clade_name")
  }

  # Samples by treatment
  treatment_samples <- list()
  for (trt in treatments) {
    samp <- metadata %>%
      dplyr::filter(!!rlang::sym(treatment_col) == trt) %>%
      dplyr::pull(!!rlang::sym(sample_col))
    treatment_samples[[trt]] <- unique(as.character(samp))
  }

  # Prepare table (NA -> 0) only in present sample columns
  all_sample_cols <- unlist(treatment_samples)
  abundance_means <- abundance_data %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(all_sample_cols[all_sample_cols %in% colnames(abundance_data)]),
        ~ as.numeric(tidyr::replace_na(.x, 0))
      )
    )

 # Means by treatment
  for (trt in treatments) {
    cols <- treatment_samples[[trt]]
    cols <- cols[cols %in% colnames(abundance_means)]
    if (length(cols) > 0) {
      abundance_means[[trt]] <- rowMeans(dplyr::select(abundance_means, dplyr::all_of(cols)), na.rm = TRUE)
    } else {
      abundance_means[[trt]] <- 0
    }
  }

# Lists of present taxa
  taxa_lists <- list()
  for (trt in treatments) {
    taxa_lists[[trt]] <- abundance_means %>%
      dplyr::filter(!!rlang::sym(trt) > presence_threshold) %>%
      dplyr::pull(clade_name)
  }

  list(taxa_lists = taxa_lists, treatments = treatments)
}

# ---- (2) FUNCTION: generate and save proportional Euler for a pair
make_euler_for_pair <- function(dos_letras,
                                datos_est,
                                abund_t,
                                out_dir = "graficos/parejas/venn_eulerr",
                                presence_threshold = 0,
                                sample_col = "Nombre") {

  # Filer
  datos_filtrados <- datos_est %>%
    dplyr::mutate(Componentes = as.character(Componentes)) %>%
    dplyr::filter(grepl(paste0("^[", dos_letras, "]+$"), Componentes)) %>%
    dplyr::filter(Tipo != "trío")

# Order levels: single1, pair, single2
  letters_order <- strsplit(dos_letras, "")[[1]]
  singles <- unique(datos_filtrados$Componentes)
  singles <- singles[nchar(singles) == 1]
  singles_ordered <- intersect(letters_order, singles)

  if (length(singles_ordered) != 2) {
    message("Saltando ", dos_letras, ": faltan singles (presentes = ", paste(singles, collapse = ","), ")")
    return(invisible(NULL))
  }

  niveles_deseados <- c(singles_ordered[1], dos_letras, singles_ordered[2])

  metadata_mod <- datos_filtrados %>%
    dplyr::mutate(Componentes = factor(as.character(Componentes), levels = niveles_deseados))

  # Extract taxa lists
  result <- extract_taxa_lists(
    abundance_data = abund_t,
    metadata = metadata_mod,
    treatment_col = "Componentes",
    sample_col = sample_col,
    center_treatment = dos_letras,
    presence_threshold = presence_threshold
  )

  # Proportional Euler (from lists)
  fit <- eulerr::euler(result$taxa_lists)

 # Save PNG
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  out_file <- file.path(out_dir, paste0("euler_", dos_letras, ".png"))

  ragg::agg_png(filename = out_file, width = 1200, height = 1200, res = 200)


print(plot(
  fit,
  fills = c("#d95f02", "purple", "#1b9e77"),
  alpha = 0.5,
  labels = list(font = 2, cex = 1.3),
  quantities = list(font = 2, cex = 1.3),
  main = list(label = "ASVs compartidos", font = 2, cex = 1.6)
))

  grDevices::dev.off()

  message("✓ Euler guardado: ", out_file)
  invisible(out_file)
}

# ---- (3) EXECUTION: all pairs
letras_pool <- c("A", "C", "D", "E", "F", "G", "H")
combinaciones <- gtools::combinations(n = length(letras_pool), r = 2, v = letras_pool)
combinaciones <- apply(combinaciones, 1, paste0, collapse = "")

out_files <- lapply(
  combinaciones,
  make_euler_for_pair,
  datos_est = datos_est,
  abund_t = abund_t,
  out_dir = "graficos/parejas/venn_eulerr",
  presence_threshold = 0,
  sample_col = "Nombre"
)
```
---

## Panel preparation

For each combination of two communities and their pair, a panel is created with all the previous plots. The panel title includes the two individual communities and the community that wins the matchup.

```{r}

make_panel_pair_6 <- function(
  pareja,
  titulo = NULL,
  bray_nmds_path    = paste0("graficos/parejas/Bray-Curtis/NMDS_", pareja, ".png"),
  unifrac_nmds_path = paste0("graficos/parejas/Unifrac/Weighted_Unifrac_NMDS_", pareja, ".png"),
  bray_box_path     = paste0("graficos/parejas/cajas_BrayCurtis/Boxplot_", pareja, ".png"),
  unifrac_box_path  = paste0("graficos/parejas/cajas_unifrac/Boxplot_Unifrac_", pareja, ".png"),
  euler_path        = paste0("graficos/parejas/venn_eulerr/euler_", pareja, ".png"),
  rainbow_dir       = "graficos/parejas/rainbow",
  out_path          = paste0("graficos/parejas/paneles_6pruebas/Panel_6_", pareja, ".png")
) {
  message("Procesando pareja: ", pareja)

  
  dir.create(dirname(out_path), recursive = TRUE, showWarnings = FALSE)

  find_rainbow_file <- function(pareja, rainbow_dir = "graficos/parejas/rainbow") {
  pat_mid <- paste0("rainbow_.*_", pareja, "_.*\\.png$")
  files_mid <- list.files(rainbow_dir, pattern = pat_mid, full.names = TRUE)
  if (length(files_mid) >= 1) return(files_mid[1])

  a <- substr(pareja, 1, 1); b <- substr(pareja, 2, 2)
  pat_any <- paste0("rainbow_.*", a, ".*", b, ".*\\.png$")
  files_any <- list.files(rainbow_dir, pattern = pat_any, full.names = TRUE)
  if (length(files_any) >= 1) return(files_any[1])

  NA_character_
}
  rainbow_path <- find_rainbow_file(pareja, rainbow_dir = rainbow_dir)

  paths <- c(
    bray_nmds_path, unifrac_nmds_path,
    rainbow_path,
    bray_box_path, unifrac_box_path,
    euler_path
  )
  names(paths) <- c("bray_nmds", "unifrac_nmds", "rainbow", "bray_box", "unifrac_box", "euler")

  if (any(is.na(paths)) || any(!file.exists(paths[!is.na(paths)]))) {
    missing <- paths[is.na(paths) | !file.exists(paths)]
    message("Saltando ", pareja, " (faltan archivos):\n  - ", paste(missing, collapse = "\n  - "))
    return(invisible(NULL))
  }

  read_and_fit <- function(path, size = "1200x900") {
    img <- image_read(path)
    img <- image_background(img, "white")
    img <- image_resize(img, paste0(size, ">"))
    img <- image_border(img, "white", "20x20")
    img
  }

  label_outside <- function(img, txt, band_h = 60) {
    w <- image_info(img)$width
    band <- image_blank(width = w, height = band_h, color = "white")
    band <- image_annotate(
      band, txt, gravity = "northwest",
      size = 45, color = "black", weight = 700, location = "+20+10"
    )
    image_append(c(band, img), stack = TRUE)
  }

  read_and_fit_stretchY <- function(path, size = "1200x900", stretchY = 1.15) {
    img <- image_read(path)
    img <- image_background(img, "white")
    img <- image_resize(img, paste0(size, ">"))

    info <- image_info(img)
    new_w <- info$width
    new_h <- round(info$height * stretchY)

    img <- image_resize(img, paste0(new_w, "x", new_h, "!"))
    img <- image_extent(img, size, gravity = "center", color = "white")
    img <- image_border(img, "white", "20x20")
    img
  }
  
  bray_nmds    <- read_and_fit(bray_nmds_path,    "1200x900")
  unifrac_nmds <- read_and_fit(unifrac_nmds_path, "1200x900")
  bray_box    <- read_and_fit(bray_box_path,    "1200x900")
  unifrac_box <- read_and_fit(unifrac_box_path, "1200x900")
  rainbow <- read_and_fit(rainbow_path, "1200x900")
  
  euler <- image_read(euler_path)
  euler <- image_background(euler, "white")
  euler <- image_resize(euler, "850x850")
  euler <- image_extent(euler, "1200x900", gravity = "center", color = "white")
  euler <- image_border(euler, "white", "20x20")

  bray_nmds    <- label_outside(bray_nmds,    "A)")
  unifrac_nmds <- label_outside(unifrac_nmds, "B)")
  bray_box     <- label_outside(bray_box,     "C)")
  unifrac_box  <- label_outside(unifrac_box,  "D)")
  euler        <- label_outside(euler,        "E)")
  rainbow      <- label_outside(rainbow,      "F)")
  
  #   top    <- image_append(c(bray_nmds, unifrac_nmds, euler))
  # bottom <- image_append(c(bray_box, unifrac_box, rainbow))
  # panel  <- image_append(c(top, bottom), stack = TRUE)

row1 <- image_append(c(bray_nmds,    unifrac_nmds))   # A | B
row2 <- image_append(c(bray_box,     unifrac_box))    # C | D
row3 <- image_append(c(euler,        rainbow))        # E | F

panel <- image_append(c(row1, row2, row3), stack = TRUE)

  panel <- image_border(panel, "white", "0x110")
  panel <- image_annotate(
    panel, titulo,
    gravity = "north",
    size = 60, color = "black", weight = 700
  )

  image_write(panel, out_path, format = "png")
  message("✓ Panel creado: ", out_path)
  invisible(out_path)
}

#Adjust according to the results obtained
titulos <- c(
  "AC" = "Comparación A vs C: ganador A",
  "AD" = "Comparación A vs D: ganador D",
  "AE" = "Comparación A vs E: ganador E",
  "AF" = "Comparación A vs F: ganador F",
  "AG" = "Comparación A vs G: ganador A",
  "AH" = "Comparación A vs H: no hay ganador",
  "CD" = "Comparación C vs D: ganador D",
  "CE" = "Comparación C vs E: ganador E",
  "CF" = "Comparación C vs F: ganador F",
  "CG" = "Comparación C vs G:no hay ganador",
  "CH" = "Comparación C vs H: ganador H",
  "DE" = "Comparación D vs E: ganador E",
  "DF" = "Comparación D vs F: ganador D",
  "DG" = "Comparación D vs G: ganador D",
  "DH" = "Comparación D vs H: no hay ganador",
  "EF" = "Comparación E vs F: ganador E",
  "EG" = "Comparación E vs G: ganador E",
  "EH" = "Comparación E vs H: ganador H",
  "FG" = "Comparación F vs G: ganador F",
  "FH" = "Comparación F vs H: no hay ganador",
  "GH" = "Comparación G vs H: ganador H"
)


Suelos <- c("A","C","D","E","F","G","H")
pares <- combn(Suelos, 2, simplify = FALSE)
parejas <- vapply(pares, \(x) paste0(x[1], x[2]), character(1))

out_files_6pruebas <- lapply(parejas, function(p) {
  make_panel_pair_6(
    pareja = p,
    titulo = titulos[[p]]  
  )
})
```
---

## Results network 

Using the results of the matchups between the individual communities, a network of combinations is generated showing the winner of each matchup.

```{r}

txt <- c(
  "A gana a C",
  "D gana a A",
  "E gana a A",
  "F gana a A",
  "A gana a G",
  "A empata con H",
  "D gana a C",
  "E gana a C",
  "F gana a C",
  "C empata con G",
  "H gana a C",
  "E gana a D",
  "D gana a F",
  "D gana a G",
  "D empata con H",
  "E gana a F",
  "E gana a G",
  "E gana a H",
  "F gana a G",
  "F empata con H",
  "H gana a G"
)

# Parse to edge data frame
edges <- lapply(txt, function(x) {
  if (str_detect(x, " defeats ")) {
    m <- str_match(x, "^\\s*([A-Z])\\s+gana a\\s+([A-Z])\\s*$")
    data.frame(from = m[2], to = m[3], Resultado = "win", stringsAsFactors = FALSE)
  } else if (str_detect(x, " draft with ")) {
    m <- str_match(x, "^\\s*([A-Z])\\s+draws  with\\s+([A-Z])\\s*$")
    # tie = two directionss
    rbind(
      data.frame(from = m[2], to = m[3], Resultado = "Empate", stringsAsFactors = FALSE),
      data.frame(from = m[3], to = m[2], Resultado = "Empate", stringsAsFactors = FALSE)
    )
  } else {
    stop(paste("No entiendo esta línea:", x))
  }
}) |> bind_rows()

edges


g <- graph_from_data_frame(edges, directed = TRUE)

tg <- as_tbl_graph(g)

set.seed(1)
red_parejas <- ggraph(tg, layout = "fr") +
  geom_edge_link(
    aes(color = Resultado, linetype = Resultado),
    arrow = arrow(length = unit(3, "mm")),
    end_cap = circle(4, "mm"),
    edge_width = 0.8,
    alpha = 0.9
  ) +
  geom_node_point(size = 8, color = "white") +
  geom_node_text(aes(label = name), size = 6) +
  scale_edge_color_manual(values = c(Gana = "steelblue3", Empate = "gray60")) +
  scale_edge_linetype_manual(values = c(Gana = "solid", Empate = "solid")) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 14),
    plot.title = element_text(size = 20, face = "bold")
  ) +
  ggtitle("Red dirigida de resultados de coalescencia")

ggsave("graficos/red_resultados.png", plot = red_parejas, width = 10, height = 5, dpi = 300, bg ="white")


```


